---
title: "Rice's Whale Vessel Risk Analysis by Owner in Gulf of America"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Overview

```{r setup}
# Install and load required packages
librarian::shelf(
  dplyr, duckdb, duckdbfs, ggplot2, glue, here, janitor, leafgl, leaflet, leaflet.extras, mapview,
  readr, sf, stringr, terra, units, viridis)

mapviewOptions(
  basemaps       = "Esri.OceanBasemap",
  vector.palette = \(n) grDevices::hcl.colors(n, palette = "Spectral", rev = T),
  raster.palette = \(n) grDevices::hcl.colors(n, palette = "Spectral", rev = T) )

options(readr.show_col_types = F)

dir_data      <- "~/My Drive/projects/ricei/data"
ship_cell_bq  <- "benioff-ocean-initiative.gulf_of_mexico_2023.vessel_presence_bymmsi_10_v1"
ship_bq       <- "benioff-ocean-initiative.gulf_of_mexico_2023.ihs_data_all"
nmfs_gdb      <- "~/My Drive/projects/msens/data/raw/fisheries.noaa.gov/NMFS_ESA_Critical_Habitat_20230505.gdb"
crithab_geo   <- here("data/nmfs_crit_hab.geojson")
subareas_shp  <- glue("{dir_data}/raw/Rice's Whale Proposed Critical Habitat Sub-Areas/Rice\'s Whale Proposed Critical Habitat Sub-Areas.shp")
subareas_geo  <- here("data/crithab_subareas.geojson")
dir_tbl       <- glue("{dir_data}/raw/ships/gulf_of_mexico_2023")
companies_csv <- glue("{dir_tbl}/companies.csv")
gfw_db        <- "/Users/bbest/big-files_gdrive-unsync/gfw.duckdb" # global Global Fishing Watch: vessel_presence_bymmsi_10_v1
goa_db        <- glue("{dir_tbl}/goa.duckdb")  # global -> gulf of america
whales_geo    <- here("data/whales.geojson") # hex_id:  whale densities by hexagon
whales_csv    <- here("data/whales.csv")
goa_geo       <- here("data/goa.geojson")

# Connect to DuckDB
con_gfw <- dbConnect(duckdb(), dbdir = gfw_db, read_only = F)
con_goa <- dbConnect(duckdb(), dbdir = goa_db, read_only = F)
load_spatial(con_gfw)
load_spatial(con_goa)

# dbListTables(con_gfw)
# dbListTables(con_goa)
# "cell"      "company"   "goa"       "ship"      "ship_cell"
# v_lon <- tbl(con_goa, "cell") |> distinct(cell_ll_lon) |> arrange(cell_ll_lon) |> pull(cell_ll_lon) 
# diff(v_lon) |> min() # 0.1

```

## Read Whale Density Hexagons

```{r read_whale_densities}
h_whales <- read_sf(whales_geo) |> # [h]exagons
  st_make_valid() 
d_whales <- read_csv(whales_csv)   # [d]ata

# h_whales
# Simple feature collection with 13296 features and 2 fields
# Geometry type: POLYGON
# Dimension:     XY
# Bounding box:  xmin: -97.02703 ymin: 23.83333 xmax: -80.45734 ymax: 30.17497
# Geodetic CRS:  WGS 84
# # A tibble: 13,296 × 3
#    hex_id area_km2                        geometry
#     <dbl>    <dbl>                   <POLYGON [°]>
#  1 136253     39.5 ((-86.93031 30.09746, -86.9586…
#  2 136254     39.5 ((-86.7977 30.08583, -86.83793…
#  3 136255     39.5 ((-86.67701 30.0751, -86.71723…
#  4 136256     39.5 ((-86.55635 30.06422, -86.5965…
#  5 136257     39.5 ((-86.43572 30.05321, -86.4759…
#  6 136258     39.5 ((-86.31513 30.04205, -86.3553…
#  7 136646     39.5 ((-86.93031 30.03624, -86.9656…
#  8 136647     39.5 ((-86.80477 30.02523, -86.8449…
#  9 136648     39.5 ((-86.68417 30.0145, -86.72437…
# 10 136649     39.5 ((-86.56361 30.00364, -86.6037…
#
# d_whales
# A tibble: 518,544 × 5
#    hex_id timespan month metric           value
#     <dbl> <chr>    <chr> <chr>            <dbl>
#  1 136253 avg      NA    cv        13.7        
#  2 136253 avg      NA    n_per_km2  0.00000125 
#  3 136253 avg      NA    se         0.00158    
#  4 136254 avg      NA    cv        12.3        
#  5 136254 avg      NA    n_per_km2  0.00000201 
#  6 136254 avg      NA    se         0.00279    
#  7 136255 avg      NA    cv        14.7        
#  8 136255 avg      NA    n_per_km2  0.000000993
#  9 136255 avg      NA    se         0.00148    
# 10 136256 avg      NA    cv        14.9 
# 
# d_whales$timespan |> table()
#    avg  month 
#  39888 478656
#
# d_whales$month |> table()
#   apr   aug   dec   feb   jan   jul   jun   mar   may   nov   oct   sep 
# 39888 39888 39888 39888 39888 39888 39888 39888 39888 39888 39888 39888

h_w_n <- h_whales |> 
  left_join(
    d_whales |> 
      filter(
        timespan == "avg",
        metric   == "n_per_km2") |> 
      select(hex_id, value) |> 
      mutate(
        n_per_1000km2 = value * 1000),
    by = "hex_id") |> 
  relocate(n_per_1000km2)

cols <- makeColorMatrix(~n_per_1000km2, h_w_n, "Spectral", reverse = T)

pal <- colorNumeric(
  palette = "Spectral",
  reverse = T,
  domain  = h_w_n$n_per_1000km2)

leaflet() |>
  addProviderTiles(provider = providers$Esri.OceanBasemap) |>
  addGlPolygons(data = h_w_n, color = cols, popup = TRUE) |> 
  addLegend(
    pal    = pal, 
    values = range(h_w_n$n_per_1000km2), 
    title  = "whales per<br>1,000 km^2")
```

## Read Critical Habitat Polygon

```{r read_crithab}
# Read NMFS Critical Habitat

if (!file.exists(crithab_geo)) {
  message("Reading NMFS Critical Habitat from GDB...")
  
  # st_layers(nmfs_gdb) |> 
  #   tibble() |> 
  #   select(name, features, fields) |> 
  #   filter(str_detect(name, "Rice"))
  #   name                         features fields
  #   <chr>                           <dbl>  <dbl>
  # 1 Proposed_RicesWhale_20230724        1     20  

  ply_crithab <- read_sf(nmfs_gdb, layer = "Proposed_RicesWhale_20230724") |> 
    clean_names() |> 
    select(-shape_area, -shape_length) |> 
    st_set_crs(4326) |>
    mutate(
      area_km2 = st_area(Shape) |> set_units("km^2"))
  st_geometry(ply_crithab) <- "geometry"
  
  write_sf(ply_crithab, crithab_geo, delete_dsn = TRUE)
}
ply_crithab <- read_sf(crithab_geo)

mapView(
  ply_crithab, 
  col.regions = "red",
  layer.name = "critical habitat") %>%
  .@map |> 
  addFullscreenControl()
```

## Read Subarea Polygons

```{r read_subareas}
# Read Rice's Whale Proposed Critical Habitat Sub-Areas
if (!file.exists(subareas_geo)) {
  message("Reading Rice's Whale Proposed Critical Habitat Sub-Areas from Shapefile...")
  
  ply_subareas <- read_sf(subareas_shp) |> 
    clean_names() |> 
    select(-shape_area, -shape_leng) |> 
    rename(subarea_id = sub_area_i) |>
    st_set_crs(4326) |>
    mutate(
      area_km2 = st_area(geometry) |> set_units("km^2"))
  st_geometry(ply_crithab) <- "geometry"
  
  write_sf(ply_subareas, subareas_geo, delete_dsn = TRUE)
}
ply_subareas <- read_sf(subareas_geo)
pts_subareas <- ply_subareas |> 
  mutate(geometry = st_centroid(geometry))

mapView(
  ply_subareas, 
  zcol       = "subarea_id",
  layer.name = "subarea") %>%
  .@map |> 
  addFullscreenControl() |> 
  addLabelOnlyMarkers(
    data  = pts_subareas,
    label = ~as.character(subarea_id), 
    labelOptions = labelOptions(
      noHide    = T,
      direction = 'top',
      textOnly  = T))
```

## Read Ship Cells

- source: [benioff-ocean-initiative.gulf_of_mexico_2023.vessel_presence_bymmsi_10_v1](https://console.cloud.google.com/bigquery?ws=!1m4!1m3!3m2!1sbenioff-ocean-initiative!2sgulf_of_mexico_2023)

```{r bigquery_to_parquet}
#| eval: false

librarian::shelf(
  arrow, bigrquery)

# old: run in bigquery ----

# Define project and dataset information
# project_id <- "benioff-ocean-initiative"
# dataset_id <- "gulf_of_mexico_2023"
# table_id   <- "vessel_presence_bymmsi_10_v1"

# Define SQL query
# Note: Using partition filtering on date for efficiency
# sql <- glue("
#   SELECT 
#     cell_ll_lat,
#     cell_ll_lon,
#     dist_travelled_km,
#     avg_speed_knots,
#     CASE 
#       WHEN avg_speed_knots <= 10 THEN '0-10'
#       WHEN avg_speed_knots <= 12 THEN '10-12'
#       WHEN avg_speed_knots <= 15 THEN '12-15'
#       ELSE '15+'
#     END AS speed_bin
#   FROM `{project_id}.{dataset_id}.{table_id}`
#   WHERE date BETWEEN '2023-01-01' AND '2023-12-31'
#     AND cell_ll_lat IS NOT NULL
#     AND cell_ll_lon IS NOT NULL
#     AND dist_travelled_km IS NOT NULL
#     AND avg_speed_knots IS NOT NULL")

# data <- bq_table_download(bq_project_query(project_id, sql))
# Error in `bq_post()`:
# ! Access Denied: Project benioff-ocean-initiative: User does not have
#   bigquery.jobs.create permission in project benioff-ocean-initiative.
#   [accessDenied]
# bq_tbl <- bq_table(project_id, dataset_id, table_id)

# new: download from bigquery ----

# assumes benbest@ucsd.edu has been granted access

# mem.maxVSize()
# [1] 24576 # Mb

# Authentication - this will prompt for authentication in browser
# If you're running this in an automated environment, consider using service account authentication
bq_auth()

for (tbl_bq in c(ship_cell_bq, ship_bq)){
  # tbl_bq = ship_cell_bq # 53 chunks
  # tbl_bq = ship_bq      # 1  chunk

  message(glue("Fetching data from BigQuery: {tbl_bq} ~ {Sys.time()}"))
  p <- unlist(strsplit(tbl_bq, "[.]"))
  
  bq_tbl <- bq_table(p[1], p[2], p[3])
  
  # iterate over 1 gb chunks, given number of rows per chunk
  gb_tbl   <- as.vector(bq_table_size(bq_tbl)) / 1000^3
  nrow_tbl <- bq_table_nrow(bq_tbl)
  gb_i     <- 1                         # 1 GB chunks
  n_i      <- ceiling(gb_tbl   / gb_i)  # number of chunks;          53
  nrow_i   <- ceiling(nrow_tbl / n_i )  # rows per chunk;    16,934,451
  
  row_beg <- 0
  for (i in 1:n_i){ # i = 1
    
    pq <- glue("{dir_data}/raw/ships/{p[2]}/{p[3]}/chunk_{sprintf('%02d', i)}.parquet")
    dir.create(dirname(pq), recursive = T, showWarnings = F)
    
    if (file.exists(pq)) {
      row_beg <- row_beg + nrow_i
      next
    }
    
    message(glue("- reading chunk {i}: row {row_beg} to {min(row_beg + nrow_i, nrow_tbl)} ~ {Sys.time()}"))
    # reading chunk 1: 0 to 16934451 ~ 2025-04-14 18:16:55.603398
    # reading chunk 2: 16934451 to 33868902 ~ 2025-04-14 18:17:44.441115
    # reading chunk 36: 592705785 to 609640236 ~ 2025-04-14 21:40:13.27480
    # reading chunk 49: 812853648 to 829788099 ~ 2025-04-14 21:50:19.7238173
    data <- bq_table_download(
      bq_tbl,
      start_index = row_beg,
      n_max       = nrow_i)
    # Downloading first chunk of data.Received 127,988 rows in the first chunk.
    # Downloading the remaining 16,806,463 rows in 176 chunks of (up to) 95,991 rows.
    
    message(glue("-   writing chunk {i} to {basename(pq)} ~ {Sys.time()}"))
    # writing chunk 1 to chunk_01.parquet ~ 2025-04-14 18:17:41.160944
    write_parquet(data, pq) # compression = default_parquet_compression() = "snappy"
    
    row_beg <- row_beg + nrow_i
  }
}
```

- <https://arrow.apache.org/docs/r/articles/arrow.html#multi-file-data-sets>

TODO:

- [ ] raster of GoA vs global
- [ ] duckdb comparison of dates
  - [R Client – DuckDB](https://duckdb.org/docs/stable/clients/r.html)\
  
    ```r
    # Establish a set of Parquet files
    dbExecute(con, "COPY flights TO 'dataset' (FORMAT parquet, PARTITION_BY (year, month))")
  
    # Summarize the dataset in DuckDB to avoid reading 12 Parquet files into R's memory
    tbl(con, "read_parquet('dataset/**/*.parquet', hive_partitioning = true)") |>
      filter(month == "3") |>
      summarise(delay = mean(dep_time, na.rm = TRUE)) |>
      collect()
    ```
  - [Querying Parquet with Precision Using DuckDB – DuckDB](https://duckdb.org/2021/06/25/querying-parquet.html)\
    consider paritioning, ordering, indexing 
    - [Working with multi-file data sets • Arrow R Package](https://arrow.apache.org/docs/r/articles/dataset.html#partitioning-performance-considerations)

```{r parquet_to_duckdb}
#| eval: false

# Load Parquet files into DuckDB
dbExecute(con_gfw, glue(
  "CREATE TABLE ship_cell AS SELECT * FROM read_parquet('{dir_tbl}/vessel_presence_bymmsi_10_v1/*.parquet')") )
# nrows: 897,525,872

# dbExecute(con_gfw, "DROP TABLE cell")

# generate table cell from unique cell_ll_lon, cell_ll_lat and store permanently
tbl(con_gfw, "ship_cell") |> 
  distinct(cell_ll_lon, cell_ll_lat) |> 
  arrange(cell_ll_lon, cell_ll_lat) |> 
  mutate(cell_id = row_number()) |>
  compute("cell", temporary = F)

dbExecute(con_gfw, glue(
  "ALTER TABLE ship_cell ADD COLUMN cell_id INTEGER"))
# update ship_cell with cell_id based on matching cell_ll_lon and cell_ll_lat
dbExecute(con_gfw, glue(
  "UPDATE ship_cell AS sc
   SET cell_id = c.cell_id
   FROM cell AS c
   WHERE
     sc.cell_ll_lon = c.cell_ll_lon AND
     sc.cell_ll_lat = c.cell_ll_lat"))

# confirm min diff is 1/10th degree
# v <- tbl(con_gfw, "cell") |> 
#   filter(cell_ll_lon == -180) |> 
#   distinct(cell_ll_lat) |>
#   arrange(cell_ll_lat) |> 
#   pull(cell_ll_lat)
# range(diff(v))
# 0.1 14.7

# update table cell with geometry based on the lower-left longitude (cell_ll_lon) and latitude (cell_ll_lat)
dbExecute(con_gfw, glue(
  "ALTER TABLE cell ADD COLUMN geom GEOMETRY"))
dbExecute(con_gfw, glue(
  "UPDATE cell SET geom = ST_MakeEnvelope(
    cell_ll_lon, 
    cell_ll_lat, 
    cell_ll_lon + 0.1, 
    cell_ll_lat + 0.1)"))

dbExecute(con_gfw, glue(
  "CREATE TABLE ship AS SELECT * FROM read_parquet('{dir_tbl}/ihs_data_all/*.parquet')") )
# nrows:     236,134

# Load Parquet files into DuckDB
dbExecute(con_gfw, glue(
  "CREATE OR REPLACE TABLE company AS SELECT * FROM read_csv('{companies_csv}')") )
# nrows:          16

# Query the data
# tbl(con_gfw, "ship_cell")
# tbl(con_gfw, "ship")
#    imo_lr_ihs_no      mmsi name_of_ship         callsign ship_type group_owner technical_manager shipmanager registered_owner operator
#            <int>     <int> <chr>                <chr>    <chr>     <chr>       <chr>             <chr>       <chr>            <chr>   
#  1       8352526 244184361 TENACIOUS            PH9945   Towing/P… CFT         CFT               CFT         CFT              CFT     
#  2       9582893 244700202 AURISSE              PB7611   Chemical… CFT         CFT               CFT         CFT              CFT     
#  3       9026148 345070136 PLATTE RIVER         XCCJ8    Crew Boat CSP         CSP               CSP         HSP              CSP     
#  4       8600909 205114000 FELIX ANTOIE TSHISE… ORKO     Trailing… CVM         CVM               CVM         CVM              CVM     
#  5       9314129 210761000 SALT LAKE CITY       C4DS2    Bulk Car… Diana Ship… DWM               DWM         Marfort Navigat… Uniper …
#  6       9314129 210761000 SALT LAKE CITY       C4DS2    Bulk Car… OceanPal I… DWM               DWM         Marfort Navigat… Five Oc…
#  7       9286633 311773000 PROTEFS              C6TT6    Bulk Car… OceanPal I… DWM               DWM         Cypres Enterpri… HSL Shi…
#  8       9297929 311912000 CALIPSO              C6UH2    Bulk Car… Diana Ship… DWM               DWM         Darien Compania… Transgr…
#  9       9297929 311912000 CALIPSO              C6UH2    Bulk Car… Diana Ship… DWM               DWM         Darien Compania… Transgr…
# 10       9297929 311912000 CALIPSO              C6UH2    Bulk Car… Diana Ship… DWM               DWM         Darien Compania… Transgr…
# # ℹ more rows
# # ℹ 7 more variables: operator_code <int>, registered_owner_code <int>, chartered_from <chr>, source_table <chr>,
# #   updated_date <date>, length_m <dbl>, weight_gt <int>

tbl(con_gfw, "ship") |> 
  group_by(mmsi) |> 
  summarize(
    n = n()) |> 
  filter(n > 1) |> 
  arrange(desc(n)) |> 
  collect()
# # A tibble: 28,107 × 2
#         mmsi     n
#        <int> <dbl>
#  1 305293000    27
#  2 636015872    27
#  3 538002591    27
d_dupe1 <- tbl(con_gfw, "ship") |> 
  filter(mmsi == 305293000) |> 
  relocate(updated_date) |> 
  arrange(desc(updated_date)) |> 
  collect()
View(d_dupe1)
```

## Gulf of America

- [Marine Regions · Gulf of Mexico (IHO Sea Area)](https://www.marineregions.org/gazetteer.php?p=details&id=4288): `id=4288`

```{r fig-goa}
#| fig-cap: "Gulf of America + whale densities + 10 km buffer"

librarian::shelf(mregions2)
# sf_use_s2() # TRUE so buffer in meters

if (!file.exists(goa_geo)){
  gomex <- gaz_geometry(4288)
  
  p_whales <- h_whales |> 
    st_union()
  
  p_goa <- st_union(gomex, p_whales) |> 
    st_buffer(10000) # 10 km buffer
  
  write_sf(p_goa, goa_geo, delete_dsn = TRUE)
} else {
  p_goa <- read_sf(goa_geo)
}

(mapview(
  p_goa, 
  col.regions = "blue",
  layer.name = "Gulf of Mexico + whale densities + 10 km buffer") +
    mapview(
      gomex, 
      col.regions = "green",
      layer.name = "Gulf of Mexico") +
    mapView(
      p_whales, 
      col.regions = "red",
      layer.name = "whale densities")) %>%
  .@map |> 
  addFullscreenControl()
```

### Subset Global Data to Gulf of America

```{r subset_goa}
#| eval: false

p_goa <- read_sf(goa_geo)

librarian::shelf(duckspatial)

ddbs_write_vector(con_gfw, p_goa, "goa")

# dbExecute(con_gfw, glue(
#   "ALTER TABLE cell ADD COLUMN crs_duckspatial VARCHAR DEFAULT 'EPSG:4326'"))

# Attach global database to the GOA database connection
dbExecute(con_goa, glue("ATTACH '{gfw_db}' AS gfw"))
# dbExecute(con_goa, glue("DETACH gfw;"))

# Execute the spatial intersection query and create new table
dbExecute(
  con_goa, 
  "CREATE TABLE cell AS
    SELECT c.*
    FROM gfw.cell c
    JOIN gfw.goa g
      ON ST_Intersects(c.geom, g.geometry)") # 13,950

# Similarly write ship_cell to goa db based on cell.cell_id
dbExecute(
  con_goa, 
  "CREATE TABLE ship_cell AS
    SELECT sc.*
    FROM gfw.ship_cell sc
    JOIN cell c
      ON sc.cell_id = c.cell_id")  # 13,838,148

# convert ship_cell.mmsi to integer, so matches ship.mmsi
dbExecute(con_goa, glue(
  "ALTER TABLE ship_cell 
   ALTER COLUMN mmsi SET DATA TYPE INTEGER USING mmsi::INTEGER"))

# DB Relationships (one-to-many) ----
# cell.cell_id   -> ship_cell.cell_id
# ship_cell.mmsi -> ship.mmsi
# ship.operator  -> company.operator_lookup

# tbl(con_goa, "ship") |> count() |> collect()                        # 49,145,861: all
# tbl(con_goa, "ship_cell") |> distinct(mmsi) |> count() |> collect() #     10,542: goa distinct
# dbGetQuery(con_goa, "
#   SELECT COUNT(*) AS n
#   FROM (
#     SELECT DISTINCT mmsi
#     FROM gfw.ship_cell)")                                           #    713,985: all distinct

tbl(con_gfw, "ship") |> 
  filter(mmsi %in% (tbl(con_goa, "ship_cell") |> distinct(mmsi) |> pull())) |>
  arrange(mmsi, updated_date, source_table) |>
  tibble::rowid_to_column("ship_id") |>
  count() |> 
  collect() #     10,542: goa distinct

dbExecute(
  con_goa, 
  "CREATE OR REPLACE TABLE ship AS
    SELECT s.*
    FROM gfw.ship s
    WHERE
      s.mmsi IN (SELECT DISTINCT mmsi FROM ship_cell)")               #     32,838: goa
# ship records: gfw -> goa
# 49,145,861: gfw
#    713,985: gfw distinct
#     32,838: goa
#     10,542: goa distinct

# get ship rows with duplicate mmsi, sort by most recent updated_date
# d_dupe2 <- tbl(con_goa, "ship") |> 
#   group_by(mmsi) |> 
#   summarize(
#     n = n(),
#     updated_date = max(updated_date, na.rm = T)) |> 
#   filter(n > 1) |> 
#   arrange(desc(n)) |> 
#   collect()
# d_dupe2 |> 
#   View()
# 
# d_dupe2_top <- tbl(con_goa, "ship") |> 
#   filter(mmsi == 636015872) |> 
#   relocate(mmsi, updated_date, source_table) |>
#   arrange(updated_date) |> 
#   collect()
# View(d_dupe2_top)

# update field ship.ship_id as unique incrementing integer based on sorting by mmsi, updated_date, source_table
dbExecute(con_goa, glue(
  "ALTER TABLE ship ADD COLUMN ship_id INTEGER"))
dbExecute(con_goa, "
  UPDATE ship 
  SET ship_id = (
    SELECT row_number() OVER (
      ORDER BY mmsi, updated_date, source_table )
    FROM ship s2 
    WHERE s2.rowid = ship.rowid)")
tbl(con_goa, "ship") |> pull(rowid)

dbExecute(con_goa, glue(
  "UPDATE ship SET ship_id = ROW_NUMBER() OVER ()"))
  
dbExecute(con_goa, glue(
  "UPDATE ship SET ship_id = sub.ship_id
   FROM (
     SELECT 
       ship_id,
       mmsi,
       updated_date,
       source_table
     FROM (
       SELECT 
         ROW_NUMBER() OVER (ORDER BY mmsi, updated_date, source_table) AS ship_id,
         mmsi,
         updated_date,
         source_table
       FROM ship
     ) AS ordered_ships
   ) AS sub
   WHERE 
     ship.mmsi = sub.mmsi AND
     ship.updated_date = sub.updated_date AND
     ship.source_table = sub.source_table"))
# tbl(con_goa, "ship") |> 
#   pull(ship_id) |> 
#   duplicated() |> 
#   sum()
#   
#   filter(mmsi == 636015872) |> 
#   arrange(ship_id) |> 
#   relocate(ship_id, mmsi, updated_date, source_table) |> 
#   collect() |> 
#   View()

# copy company to goa db
dbExecute(
  con_goa, 
  "CREATE TABLE company AS
    SELECT * FROM gfw.company")  # 16

# copy goa to goa db for study area
dbExecute(
  con_goa, 
  "CREATE TABLE goa AS
    SELECT * FROM gfw.goa")  # 1

dbExecute(con_goa, glue("DETACH gfw;"))
# DBI::dbListTables(con_gfw)
# DBI::dbListTables(con_goa)
# [1] "cell"      "company"   "goa"       "ship"      "ship_cell"

dbListFields(con_goa, "ship_cell")

# retrieve and map the cells
cell <- tbl(con_goa, "cell") |> 
  mutate(geom_wkt = sql("ST_AsText(geom)")) |>
  select(-geom) |> 
  collect() |>
  st_as_sf(wkt = "geom_wkt", crs = 4326)
mapView(cell)

# extract ship_cell within Gulf of America
dbExecute(con_goa, glue(
  "CREATE TABLE ship_cell AS
    SELECT sc.*
    FROM gfw.ship_cell sc
    JOIN cell c
      ON sc.cell_id = c.cell_id;"))

dbListTables(con_goa)
dbListTables(con_gfw)
dbListFields(con_gfw, "goa")
tbl(con_gfw, "goa") |> 
  pull("crs_duckspatial") |> 
  st_transform(3857) |> 
  st_area() |> 
  sum() |> 
  set_units("km^2")

# Subset global data to Gulf of America
dbExecute(con_goa, glue(
  "CREATE TABLE ship_cell AS 
   SELECT * FROM read_parquet('{dir_tbl}/vessel_presence_bymmsi_10_v1/*.parquet')
   WHERE
     cell_ll_lon >= {st_bbox(goa_whales)[1]} AND
     cell_ll_lat >= {st_bbox(goa_whales)[2]} AND
     cell_ll_lon <= {st_bbox(goa_whales)[3]} AND
     cell_ll_lat <= {st_bbox(goa_whales)[4]}") )
```


```{r fig-time_series}
b <- st_bbox(h_whales) |> as.numeric()

# [s]hip [c]ell date number of observations
sc_date_n <- tbl(con_goa, "ship_cell") |> 
  filter(
    cell_ll_lon >= !!b[1],
    cell_ll_lat >= !!b[2],
    cell_ll_lon <= !!b[3],
    cell_ll_lat <= !!b[4]) |>
  group_by(date) |> 
  summarize(
    n = n()) |> 
  arrange(date) |> 
  collect()
# View(sc_date_n)
# surprising:
# - gaps in dates (59 days with data / 365 days)
# - spikes in data with 3 orders of magnitude diff, 
#   eg 2023-01-29 127,908 to 2023-01-31 123,249,963

librarian::shelf(dygraphs)

dygraph(sc_date_n, main = "Ship Presence in Gulf by Date") |> 
  dySeries("n", label = "Number of Records") |> 
  dyAxis("y", label = "Records") |> 
  dyRangeSelector() |> 
  dyOptions(
    strokeWidth = 1, 
    drawPoints  = T,
    pointSize   = 2)
```


```{r fig-goa_dates}
b <- st_bbox(h_whales) |> as.numeric()

# [s]hip [c]ell date number of observations
sc_date_n <- tbl(con, "ship_cell") |> 
  filter(
    cell_ll_lon >= !!b[1],
    cell_ll_lat >= !!b[2],
    cell_ll_lon <= !!b[3],
    cell_ll_lat <= !!b[4]) |>
  group_by(date) |> 
  summarize(
    n = n()) |> 
  arrange(date) |> 
  collect()
# View(sc_date_n)
# surprising:
# - gaps in dates (59 days with data / 365 days)
# - spikes in data with 3 orders of magnitude diff, 
#   eg 2023-01-29 127,908 to 2023-01-31 123,249,963

librarian::shelf(dygraphs)

dygraph(sc_date_n, main = "Ship Presence in Gulf by Date") |> 
  dySeries("n", label = "Number of Records") |> 
  dyAxis("y", label = "Records") |> 
  dyRangeSelector() |> 
  dyOptions(
    strokeWidth = 1, 
    drawPoints  = T,
    pointSize   = 2)
```

```{r company_ships}
librarian::shelf(stringr)

d_c <- tbl(con_gfw, "company") |> 
  collect()

for (i in 1:nrow(d_c)){ # i = 1
  company         <- d_c$company[i]
  operator_lookup <- d_c$operator_lookup[i]
  
  # message(company)
  # 
  # d <- dbGetQuery(
  #   con,
  #   glue("
  #        SELECT operator, operator_code FROM ship 
  #        WHERE
  #          operator ILIKE '%{company}%' ")) |> 
  #   tibble()
  # message(glue("  {company}: "), nrow(d))
  # d |> table() |> print()
  d <- dbGetQuery(
    con_gfw,
    glue("
         SELECT 
           '{company}'         AS company, 
           '{operator_lookup}' AS operator_lookup,
           operator
         FROM ship 
         WHERE
           operator ILIKE '%{operator_lookup}%' ")) |> 
    tibble()
  # message(glue("  {operator_lookup}: "), nrow(d),"\n")
  # d |> table() |> print()
}

#                                operator_code
# operator                         387 4427 28896 194433 259393 1826240 6514732
#   CMA CGM Asia Pacific Ltd         0    0     2      0      0       0       0
#   CMA CGM Asia Shipping Pte Ltd    0    0     0      0      0       3       0
#   CMA CGM Philippines Shipping     0    0     0      0      0       0       1
#   CMA CGM SA The French Line    1544    0     0   1225      0       0       0
#   Containerships-CMA CGM GmbH      0    1     0      0     13       0       0
```

```{r ship_cell_grid}
leaflet() |> 
  addProviderTiles("Esri.OceanBasemap") |> 
  addPolygons(
    data        = h_whales,
    fillColor   = "blue",
    fillOpacity = 0.5,
    color       = "black",
    weight      = 1,
    opacity     = 1,
    label       = ~glue("Hex ID: {hex_id}")) # |> 
  # setView(lng = -89, lat = 25, zoom = 5) # zoom: 5

# bounding box of whale densities
b_w <- h_whales |> 
  st_union() |> 
  st_as_sf() |> 
  mutate(
    one = 1)
# mapView(b_w)

r_10 <- terra::rast(
  st_bbox(h_whales) |> 
    st_as_sfc() |> 
    st_buffer(0.1*110*1000) |> 
    st_bbox() |> 
    ext(),
  resolution = c(0.1, 0.1)) # 10th degree resolution

r_10$cell_id <- 1:ncell(r_10)

mapView(r_10) + 
  mapView(h_whales)

r_w <- terra::rasterize(
  h_w_n, 
  r_10,
  field      = "n_per_1000km2",
  # fun        = "mean",
  # background = 0,
  touches    = T)

# DEBUG ----
# TODO: rasterize every month
mapView(r_w, zcol = "n_per_1000km2") +
  mapView(h_w_n, zcol = "n_per_1000km2")

r_s <- terra::rasterize(
  ply_subareas, 
  r_10,
  field      = "subarea_id",
  # fun        = "mean",
  # background = 0,
  touches    = T)
# r_s2 <- terra::rasterize(
#   ply_subareas, 
#   r_10,
#   field      = "subarea_id")
  # fun        = "mean",
  # background = 0,
  # touches    = T)

p_ch <- ply_crithab |> 
    mutate(one = 1)
r_ch <- terra::rasterize(
  p_ch, 
  r_10,
  field      = "one",
  # fun        = "mean",
  # background = 0,
  touches    = T)

mapView(h_w_n, zcol = "n_per_1000km2") +
  mapView(p_ch, zcol = "one") +
  mapView(ply_subareas, zcol = "subarea_id") +
  mapView(r_w, zcol = "n_per_1000km2") +
  mapView(r_s, zcol = "subarea_id") +
  mapView(r_ch, zcol = "subarea_id")

r_h <- terra::rasterize(
  h_whales |> 
    slice(1:3), 
  r_10, touches=T, cover = T, by = "hex_id")

r_h[["136253"]] |> 
  trim() |> 
  mapView() +
  mapView(h_whales |> slice(1))
values(r_h[["136253"]])


global(r_h[["136253"]], "sum", na.rm = T)
terra::colSums()
r_h[[1]] |> 
  trim()

plot(r_b)

terra::zonal()

r_10 <- r_10 |> 
  mask()
terra::()

mapView(r_10)+
  mapView(b_w)

terra::as.polygons()



# r_tif <- here("data/rasters.tif")
# r <- rast(r_tif)
# names(r)

d_cell_n <- tbl(con, "ship_cell") |> 
  filter(
    cell_ll_lon >= !!b[1],
    cell_ll_lat >= !!b[2],
    cell_ll_lon <= !!b[3],
    cell_ll_lat <= !!b[4]) |>
  group_by(cell_ll_lon, cell_ll_lat) |> 
  summarize(n = n()) |> 
  collect()  # 16,532 × 3

r_n <- d_cell_n |> 
  vect(
    geom = c("cell_ll_lon", "cell_ll_lat"),
    crs = "EPSG:4326", keepgeom = T) |>
  terra::rasterize(
    r_10,
    field = "n",
    fun   = "sum")

names(r_n) <- 

plet(log10(r_n), tiles="Esri.OceanBasemap", main = "log10(# ship<br>presence<br>records), 2023<br>(log10)")
providers$
librarian::shelf(
  tidyr)

d_w <- h_w |>
  st_drop_geometry() |> 
  pivot_longer(
    !hexid, 
    names_to  = "metric",
    values_to = "value") |> 
  filter(!is.na(value)) |> 
  separate_wider_delim(
    metric, "_", 
    names = c("month", "metric"))

d_w_avg <- d_w |> 
  group_by(
    hexid, metric) |> 
  summarize(
    value = mean(value),
    .groups = "drop")
    
d_w_all <- bind_rows(
  d_w_avg |> 
    mutate(
      timespan = "avg"),
  d_w |> 
    mutate(
      timespan = "month")) |> 
  select(hexid, timespan, month, metric, value)

r_w <- d_cell_n |> 
  vect(
    geom = c("cell_ll_lon", "cell_ll_lat"),
    crs = "EPSG:4326", keepgeom = T) |>
  terra::rasterize(
    r_10,
    field = "n",
    fun   = "sum")
    
    sf_use_s2(F)
    ply_whales <- hex_whales |> 
      select(hex_id) |> 
      filter(hex_id %in% unique(tbl_whales_avg$hex_id)) |> 
      st_intersection(
        plys_study) |> 
      group_by(
        hex_id) |> 
      summarize() |> 
      mutate(
        area_km2 = st_area(geometry) |> set_units("km^2"))
    sf_use_s2(T)
```

```{r company_cell}
d_company <- tbl(con, "company") |> 
  collect()

co <- d_company$company[1]
op <- d_company$operator_lookup[1]

tbl(con, "cell") |> 
  left_join(
    tbl(con, "ship_cell"),
    by = c("cell_ll_lon", "cell_ll_lat")) |>
```

```{r old_ship_cell_sum}

vesselpresence_bq <- "benioff-ocean-initiative.gulf_of_mexico_2023.vessel_presence_bymmsi_10_v1"

dir_vpres <- glue("{dir_data}/raw/ships/gulf_of_mexico_2023/vessel_presence_bymmsi_10_v1")
d_vpres <- open_dataset(dir_vpres)
head(d_vpres) |> collect()

# TODO: sort (date, mmsi, cell_ll_lon, cell_ll_lat) and save as a single parquet file
d_vpres_date <- d_vpres |> 
  group_by(date) |> 
  summarize(
    n                 = n(),
    dist_travelled_km = sum(dist_travelled_km, na.rm = T),
    avg_speed_knots   = mean(avg_speed_knots, na.rm = T)) |>
  arrange(date) |> 
  collect()
# d_vpres_date |> View()
#   date       cell_ll_lat cell_ll_lon mmsi      hours dist_travelled_km avg_speed_knots
#   <date>           <dbl>       <dbl> <chr>     <dbl>             <dbl>           <dbl>
# 1 2023-10-07       -42.5       -62.8 636022759 0.083              2.01           13.4 
# 2 2023-10-07       -41.3       -62.1 636022759 0.083              2.08           13.7 
# 3 2023-10-07       -38.6       144.  352002415 0.083              1.47            9.58
# 4 2023-10-07       -38.3       145.  352002415 0.083              1.93           13.1 
# 5 2023-10-07       -38.1       144.  352002415 0.083              1.18            7.66
# 6 2023-10-07       -37.5       102.  994210602 0.083              1.22            7.91

d_vpres_ll <- d_vpres |> 
  group_by(cell_ll_lon, cell_ll_lat) |> 
  summarize(
    n                 = n(),
    dist_travelled_km = sum(dist_travelled_km, na.rm = T),
    avg_speed_knots   = mean(avg_speed_knots, na.rm = T)) |>
  arrange(cell_ll_lon, cell_ll_lat) |> 
  collect()

d_vpres_ll <- d_vpres_ll |>
  
  diff(round(d_vpres_ll$cell_ll_lon, 1)) |> table()
range(d_vpres_ll$cell_ll_lon) # -180.0  179.9
range(d_vpres_ll$cell_ll_lat) # -89.0  89.8
  

# Process the data
message("Processing data...")
# Aggregate by cell and speed bin
raster_data <- data |> 
  group_by(cell_ll_lon, cell_ll_lat, speed_bin) |> 
  summarize(total_dist = sum(dist_travelled_km, na.rm = TRUE)) |> 
  ungroup()

# Get unique coordinates to determine raster dimensions
unique_lons <- sort(unique(raster_data$cell_ll_lon))
unique_lats <- sort(unique(raster_data$cell_ll_lat))

# Determine cell resolution based on data
lon_res <- min(diff(unique_lons))
lat_res <- min(diff(unique_lats))

# Create empty rasters for each speed bin
speed_bins <- c("0-10", "10-12", "12-15", "15+")
rasters <- list()

# Define raster extent
e <- extent(
  min(unique_lons) - (lon_res/2),
  max(unique_lons) + (lon_res/2),
  min(unique_lats) - (lat_res/2),
  max(unique_lats) + (lat_res/2)
)

message("Creating rasters...")
for (bin in speed_bins) {
  # Filter data for this speed bin
  bin_data <- raster_data %>% filter(speed_bin == bin)
  
  # Create an empty raster
  r <- raster(e, resolution = c(lon_res, lat_res))
  
  # Populate the raster
  for (i in 1:nrow(bin_data)) {
    cell <- cellFromXY(r, c(bin_data$cell_ll_lon[i], bin_data$cell_ll_lat[i]))
    r[cell] <- bin_data$total_dist[i]
  }
  
  # Store in list
  rasters[[bin]] <- r
}

# Create a combined raster
combined_raster <- rasters[["0-10"]] # Initialize with first raster
combined_raster[is.na(combined_raster)] <- 0 # Replace NA with 0

# Add all bins with proper names
names(combined_raster) <- "0-10 knots"

for (bin in speed_bins[-1]) {
  r <- rasters[[bin]]
  r[is.na(r)] <- 0
  combined_raster <- addLayer(combined_raster, r)
}
names(combined_raster)[2:4] <- c("10-12 knots", "12-15 knots", "15+ knots")

# Save the raster to file
writeRaster(combined_raster, "speed_binned_distances.tif", overwrite=TRUE)

# Plot the results
message("Creating visualization...")
par(mfrow=c(2,2))
for (i in 1:4) {
  plot(combined_raster[[i]], main=names(combined_raster)[i],
       col=viridis(100), zlim=c(0, max(maxValue(combined_raster))),
       axes=TRUE)
}

# Create a stack plot for all layers
plot(combined_raster, col=viridis(100))

message("Done! Results saved as 'speed_binned_distances.tif'")
```


## Disconnect from DuckDB

```{r db_disconnect}
# Disconnect from DuckDB
dbDisconnect(con_gfw, shutdown = TRUE)
dbDisconnect(con_goa, shutdown = TRUE)
```
