---
title: "Spatial analysis of ship-strike risk for Rice’s whale in the Gulf of Mexico"
author: "Benjamin D Best, PhD (ben@ecoquants.com)"
date: "2024-02-23"
format: revealjs
editor_options: 
  chunk_output_type: console
revealjs-plugins:
  - revealjs-text-resizer
---

```{r setup_chunks, include=FALSE}
knitr::opts_chunk$set(
  echo    = F,
  warning = F,
  message = F)

# packages ----
source(here::here("scripts/functions.R"))
source(here("scripts/paths.R"))
librarian::shelf(
  leaflet.extras2)

# [+ options arg in addRasterImage() for specifying pane]( https://github.com/rstudio/leaflet/pull/692#issue-650790764)
# remotes::install_github("rstudio/leaflet")

# data ----
ply_study  <- read_sf(study_geo)
ply_cells  <- read_sf(cells_geo)
ply_units  <- read_sf(units_gpkg)
ply_ships  <- read_sf(ships_geo)
ply_whales <- read_sf(whales_geo)
ply_bia    <- read_sf(bia_geo)
ply_wab    <- read_sf(wab_geo)
ply_wan    <- read_sf(wan_geo)
tbl_ships  <- read_csv(ships_csv)
tbl_whales <- read_csv(whales_csv)
lns_depth_contours <- read_sf(depth_contours_geo)
stk_rast   <- rast(rast_tif)
r_cells    <- stk_rast["cell_id"]
tbl_units <- ply_units |> 
  st_drop_geometry()
```

## Introduction {.smaller}

**Motivation**. The Rice's whale (*Balaenoptera ricei*) is endangered with an estimated population size of 51 individuals ([Garrison et al. 2020](https://www.fisheries.noaa.gov/resource/document/abundance-marine-mammals-waters-us-gulf-mexico-during-summers-2017-and-2018)). The Biological Opinion ([NOAA 2020](https://www.fisheries.noaa.gov/resource/document/biological-opinion-federally-regulated-oil-and-gas-program-activities-gulf-mexico)) that proposes vessel mitigation measures is based on a distribution model ([Roberts et al. 2016](https://doi.org/10.1038/srep22615)) that has been superseded by a newer model ([Litz et al. 2022](https://www.ncei.noaa.gov/access/metadata/landing-page/bin/iso?id=gov.noaa.nodc:0243469)) showing range expansion into the western Gulf of Mexico.

**Methods**. We adhered to similar methods as in the Biological Opinion ([NOAA 2020](https://www.fisheries.noaa.gov/resource/document/biological-opinion-federally-regulated-oil-and-gas-program-activities-gulf-mexico)) for creating a new Whale Area also based on the 100- to 400- meter isobaths to now extend to the western Gulf of Mexico, except without a 10 km buffer (since only marginal improvement from 94% to 97% containment of distribution). Risk to ship strike was based on the same AIS vessel traffic data (2014-2018).

## Whale Distributions

```{r}
r_1     <- rast(whales_roberts2016_img) |> project("epsg:3857")
group_1 <- "Whales / 100km2 (Roberts et al. 2016)"
r_2     <- stk_rast["whales_avg_n_per_100km2"] |> project("epsg:3857")
group_2 <- "Whales / 100km2 (Litz et al. 2016)"

r_vals <- unique(na.omit(c(values(r_1), values(r_2))))
pal         <- colorNumeric(
  palette = "Spectral", 
  domain  = r_vals, 
  reverse = T, na.color = "transparent")
legend_title <- "Whale density<br>(# / 100 km<sup>2</sup>)</small>"

n_arrow_img <- "https://cdn.pixabay.com/photo/2013/07/12/17/54/arrow-152596_960_720.png"

lyrs <- c(group_1, group_2)
  
m <- leaflet(
  options = leafletOptions(
    zoomControl        = T,
    attributionControl = F)) |>
  addMapPane("left",  zIndex = 0) |>
  addMapPane("right", zIndex = 0) |>
  addProviderTiles(
    "Esri.OceanBasemap",
    options = c(
      providerTileOptions(
        variant = "Ocean/World_Ocean_Base"),
      pathOptions(
        pane = "left")),
    layerId = "old") |>
  addProviderTiles(
    "Esri.OceanBasemap",
    options = c(
      providerTileOptions(
        variant = "Ocean/World_Ocean_Base"),
      pathOptions(
        pane = "right")),
    layerId = "new") |>
  addProviderTiles(
    "Esri.OceanBasemap",
    options = providerTileOptions(
      variant = "Ocean/World_Ocean_Reference")) |> 
  addControl(
    html     = "right: Litz et al. 2022",
    position = "topright")|> 
  addControl(
    html     = "left: Roberts et al. 2016",
    position = "topleft") |> 
  addRasterImage(
    r_1, project = F, opacity = 0.8,
    colors  = pal,
    group   = group_1,
    options = pathOptions(pane = "left")) |> 
  addRasterImage(
    r_2, project = F, opacity = 0.8,
    colors  = pal,
    group   = group_2,
    options = pathOptions(pane = "right")) |> 
  addSidebyside(
    layerId = "sidecontrols",
    leftId  = "old",
    rightId = "new") |> 
  addScaleBar(
    position = c("bottomleft")) |> 
  leafem::addLogo(n_arrow_img) |> 
  addLegend(
    pal      = pal, 
    values   = r_vals, 
    opacity  = 0.7, 
    title    = legend_title,
    position = "topright") |> 
  leaflet.extras::addFullscreenControl()

# add_depth_contours ----
grps  <- c("100 m contour", "400 m contour")
cols  <- c("black", "black")
wts   <- c(1, 2)
dashs <- c("5", "10")
    
m <- m |> 
  addPolylines(
    data = lns_depth_contours |>
      filter(depth_m == "100"),
    group  = grps[1],
    color  = cols[1], opacity = 1.0,
    weight = wts[1], dashArray = dashs[1],
    options = pathOptions(pane = "left")) |> 
  addPolylines(
    data = lns_depth_contours |> 
      filter(depth_m == "400"),
    group  = grps[2],
    color  = cols[2], opacity = 1.0,
    weight = wts[2], dashArray = dashs[2],
    options = pathOptions(pane = "left"))

lyrs <- c(lyrs, grps)

# add_ply_bia ----
grp <- "Bio. Imp. Area (LaBrecque 2016)"

m <- m |> 
  addPolygons(
    data = ply_bia, 
    color="pink", fill = F, 
    opacity = 1.0, weight=2,
    group = grp,
    options = pathOptions(pane = "left")) |> 
  hideGroup(grp)

lyrs <- c(lyrs, grp)

# add_ply_wab ----
grp <- "Whale Area (NOAA 2020)"

m <- m |> 
  addPolygons(
    data = ply_wab, 
    color="purple", fill = F, 
    opacity = 1.0, weight=2,
    group = grp,
    options = pathOptions(pane = "left"))

lyrs <- c(lyrs, grp)

# add_ply_wan ----
grp <- "Whale Area, New"

m <- m |> 
  addPolygons(
    data = ply_wan, 
    color="red", fill = F, 
    opacity = 1.0, weight=2,
    group = grp,
    options = pathOptions(pane = "right"))

lyrs <- c(lyrs, grp)

# add_layersControl ----
m |> 
  addLayersControl(
    overlayGroups = lyrs,
    options = layersControlOptions(collapsed = T))
```

## Whales Captured

```{r}
#| label: tbl-whales-by-area
#| tbl-cap: Table of new whale densities (Litz et al. 2022) summarized by total study area (U.S. Gulf of Mexico), previous Whale Area (NMFS 2020) and newly proposed Whale Area.

a <- read_csv(whales_n_by_area_csv)

d <- tribble(
  ~Item, ~`#`, ~`%`,
  "Whales in Study (U.S. Gulf of Mexico)", a$n_whales_gom, 1,
  "Whales in Original Whale Area (NMFS, 2020)", a$n_whales_wab, a$pct_whales_wab,
  "Whales in New Whale Area", a$n_whales_wan, a$pct_whales_wan)

d |> 
  gt() |> 
  fmt_number(decimals=0) |>  # c(risk_all, risk_boem), 
  fmt_percent(
    `%`, decimals = 0) |>
  opt_stylize(style = 6) |> 
  tab_options(table.font.size = px(24))
```

## Vessel Traffic {.smaller}

::: {.panel-tabset}

### All Types, All Speeds

```{r}
#| label: fig-ships-avg-all-gt01
#| fig-cap: "Map of annual average traffic (km) for all vessel types at all speeds from AIS data (2014 to 2018). Depth contours are shown in dash blacked lines for 100 m (finer) and 400 m (thicker)."

map_rast_jenks(stk_rast["ships_avg_all_gt01"], "Traffic,<br><small>all types,<br>all speeds,<br>avg year<br>(km)</small>", "RdYlGn")
```

### Oil & Gas, All Speeds

```{r}
#| label: fig-ships-avg-boem-gt01
#| fig-cap: "Map of annual average traffic (km) for oil and gas vessels at all speeds from AIS data (2014 to 2018). Depth contours are shown in dash blacked lines for 100 m (finer) and 400 m (thicker)."

map_rast_jenks(stk_rast["ships_avg_boem_gt01"], "Traffic,<br><small>oil & gas,<br>all speeds,<br>avg year<br>(km)</small>", "RdYlGn")
```

### All Types, > 10 knots

```{r}
#| label: fig-ships-avg-all-gt10
#| fig-cap: "Map of annual average traffic (km) for all vessel types > 10 knots from AIS data (2014 to 2018). Depth contours are shown in dash blacked lines for 100 m (finer) and 400 m (thicker)."

map_rast_jenks(stk_rast["ships_avg_all_gt01"], "Traffic,<br><small>all types,<br>> 10 knots,<br>avg year<br>(km)</small>", "RdYlGn")
```

### Oil & Gas, > 10 knots

```{r}
#| label: fig-ships-avg-boem-gt10
#| fig-cap: "Map of annual average traffic (km) for oil and gas vessels > 10 knots from AIS data (2014 to 2018). Depth contours are shown in dash blacked lines for 100 m (finer) and 400 m (thicker)."

map_rast_jenks(stk_rast["ships_avg_boem_gt01"], "Traffic,<br><small>oil & gas,<br>> 10 knots,<br>avg year<br>(km)</small>", "RdYlGn")
```

:::

## Vessel Risk to Whales {.smaller}

$$
Risk = {Whale\:density}\:(\#/100km^2) * {Vessel\:traffic\:(km/100km^2)}
$$

::: {.panel-tabset}

### All Types, All Speeds

```{r}
#| label: fig-risk-avg-all-gt01
#| fig-cap: "Map of risk (# whales * km vessel traffic) for all vessels at all speeds. Depth contours are shown in dash blacked lines for 100 m (finer) and 400 m (thicker)."

stk_rast["risk_avg_all_gt01"] |> 
  map_rast("Risk,<br><small>all types,<br>all speeds<br>(# * km)</small>", "RdYlGn")
```

### Oil & Gas, > 10 knots

```{r}
#| label: fig-risk-avg-boem-gt10
#| fig-cap: "Map of risk (# whales * km vessel traffic) for oil and gas vessels > 10 knots. Depth contours are shown in dash blacked lines for 100 m (finer) and 400 m (thicker)."

stk_rast["risk_avg_boem_gt10"] |> 
  map_rast("Risk,<br><small>oil & gas,<br>> 10 knots <br>(# * km)</small>", "RdYlGn")
```

:::

## Risk Reduction by Areas {.smaller}

```{r}
#| label: tbl-risk-reduction-by-areas
#| tbl-cap: "Reduction of vessel strike risk (# whales * km vessel traffic) to Rice's whales with enforcement of original (NMFS 2020) and new Whale Areas, averaged across years of AIS data (2014 to 2018). All percentage (%) reductions are compared to All Vessels for given speeds."

d <- read_csv(risk_reduction_by_areas_csv) |> 
  select(yr, matches("wan|wab$")) |> 
  rename_with(
    ~ str_replace(.x, "^pct_risk_reduced", "pctriskreduced"),
    starts_with("pct_risk_reduced")) |> 
  pivot_longer(
    -yr,
    names_to  = "var",
    values_to = "val") |>  # table(d$var)
  separate_wider_delim(
    var,
    delim = "_",
    names = c("v","type","speed","area")) |> 
  pivot_wider(
    names_from = c(v, area),
    names_sep = "_",
    values_from = val) |> 
  relocate(speed, type) |> 
  arrange(speed, type, yr)
# names(d) |> sort() |> paste(collapse="\n") |> cat()

d <- d |> 
  mutate(
    yr = as.character(yr)) |> 
  bind_rows(
    bind_cols(
      yr = "Avg",
      d |>
        select(-yr) |> 
        group_by(speed, type) |> 
        summarise(
          across(where(is.numeric), mean),
          .groups = "drop") ) ) |>
  arrange(speed, type, yr)

d <- d |> 
  mutate(
    speed = case_match(
      speed, 
      "gt01" ~ "All speeds",
      "gt10" ~ "> 10 knots"),
    type = case_match(
      type, 
      "all"  ~ "All vessels",
      "boem" ~ "Oil & Gas vessels") )

d |> 
  filter(yr == "Avg") |> 
  mutate(
    speed_type = glue("{speed} - {type}")) |> 
  select(speed_type, risk_wab, pctriskreduced_wab, risk_wan, pctriskreduced_wan) |> 
  gt() |> 
  fmt_number(decimals=0) |>  # c(risk_all, risk_boem), 
  fmt_percent(c(
    pctriskreduced_wab,
    pctriskreduced_wan), decimals = 0) |>
  opt_stylize(style = 6) |>
  cols_label(
    speed_type        = "Speed - Vessel Type",
    risk_wab           = "Original",
    pctriskreduced_wab = "%",
    risk_wan           = "New",
    pctriskreduced_wan = "%") |> 
  tab_spanner(
    label = "Risk Reduction by Whale Area",
    columns = c(
      risk_wab, pctriskreduced_wab, 
      risk_wan, pctriskreduced_wan)) |> 
  tab_options(table.font.size = px(24))
```

## Thank You

This report received funding from Earthjustice and NRDC. The AIS data used in (NMFS 2020) were provided by NMFS pursuant to a Freedom of Information Act request. A portion of the background section was contributed by Jay Barlow. Steve Mashuda (Earthjustice) and Michael Jasny (NRDC) reviewed drafts of the report.

### Further Information

- Report: [ecoquants.com/ricei](https://ecoquants.com/ricei/)
- Source code: [github.com/ecoquants/ricei](https://github.com/ecoquants/ricei)