---
title: "Rice's Whale Ship-Strike Risk Analysis"
subtitle: "NRDC Contract - 2023 AIS Data Analysis"
author: "Ben Best, EcoQuants LLC"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
---

```{r setup}
#| label: setup

# load libraries ----
librarian::shelf(
  DBI, dplyr, duckdb, glue, here, knitr, leaflet, lubridate,
  purrr, sf, tibble, tidyr, viridisLite,
  quiet = TRUE
)

# source risk functions ----
source(here("scripts/risk_functions.R"))

# set paths ----
is_server   <- Sys.info()[["sysname"]] == "Linux"
dir_private <- ifelse(is_server, "/share/private", "~/My Drive/private")
dir_data    <- ifelse(is_server, "/share/data/ricei", "~/My Drive/projects/ricei/data")
goa_db      <- glue("{dir_data}/raw/ships/gulf_of_mexico_2023/goa.duckdb")

# connect to database ----
con <- DBI::dbConnect(duckdb::duckdb(), dbdir = goa_db, read_only = TRUE)
DBI::dbExecute(con, "INSTALL spatial; LOAD spatial;")
```

## Overview

This analysis calculates ship-strike risk for Rice's whale (*Balaenoptera ricei*) in the Gulf of Mexico, using:

- **Whale densities:** Litz et al. (2022) hexagonal grid model
- **Vessel traffic:** 2023 AIS data from Benioff Ocean Science Laboratory
- **Risk model:** Conn & Silber (2013) speed-based risk factors

### Risk Factor Model

Based on Conn & Silber (2013), strike risk increases non-linearly with vessel speed:
```{r}
#| label: risk-factors-table

tibble(
  `Speed Bin`   = RISK_FACTORS$labels,
  `Risk Factor` = RISK_FACTORS$factors,
  `Rationale`   = c(
    "Baseline (slowdown target)",
    "Midpoint 11 kt",
    "Midpoint 13.5 kt",
    "Conservative at 15 kt"
  )
) |>
  kable()
```

## Data Inventory

### Database Tables

```{r}
#| label: db-inventory

# list tables
tables <- DBI::dbListTables(con)
cat("Tables in goa.duckdb:", paste(tables, collapse = ", "), "\n\n")

# count records
for (tbl_name in tables) {
  n <- tbl(con, tbl_name) |> count() |> collect() |> pull(n)
  cat(glue("{tbl_name}: {format(n, big.mark=',')} records"), "\n")
}
```

### Spatial Data Files

```{r}
#| label: spatial-inventory

dir_local_data <- here("data")

spatial_files <- c(
  "whales.geojson",
  "crithab_subareas.geojson",
  "nmfs_crit_hab.geojson",
  "goa.geojson"
)

for (f in spatial_files) {
  fpath <- file.path(dir_local_data, f)
  if (file.exists(fpath)) {
    sf_obj <- sf::st_read(fpath, quiet = TRUE)
    cat(glue("{f}: {nrow(sf_obj)} features, {ncol(sf_obj)} fields"), "\n")
  } else {
    cat(glue("{f}: NOT FOUND"), "\n")
  }
}
```

## Phase 1: Data Preparation

### 1.1 Load Spatial Boundaries

```{r}
#| label: load-spatial

# study area
ply_goa <- sf::st_read(here("data/goa.geojson"), quiet = TRUE)

# critical habitat
ply_crithab <- sf::st_read(here("data/nmfs_crit_hab.geojson"), quiet = TRUE)

# subareas (7 fairway/port approaches)
ply_subareas <- sf::st_read(here("data/crithab_subareas.geojson"), quiet = TRUE)

# whale densities
ply_whales <- sf::st_read(here("data/whales.geojson"), quiet = TRUE)
```

### 1.2 Speed Distribution Summary

```{r}
#| label: speed-distribution

# query speed distribution with bins
speed_sql <- glue("
  SELECT
    {get_speed_bin_sql()} AS speed_bin,
    COUNT(*) AS n_records,
    SUM(dist_travelled_km) AS total_km,
    SUM(hours) AS total_hours
  FROM ship_cell
  WHERE avg_speed_knots IS NOT NULL
  GROUP BY speed_bin
  ORDER BY speed_bin
")

speed_dist <- DBI::dbGetQuery(con, speed_sql)
speed_dist |>
  mutate(
    pct_records = n_records / sum(n_records) * 100,
    pct_km      = total_km / sum(total_km) * 100,
    pct_hours   = total_hours / sum(total_hours) * 100
  ) |>
  kable(
    digits = 1,
    format.args = list(big.mark = ","),
    col.names = c(
      "Speed Bin", "Records", "Distance (km)", "Hours",
      "% Records", "% Distance", "% Hours"
    )
  )
```

### 1.3 Company Fleet Summary

```{r}
#| label: company-summary

# query company traffic in critical habitat
company_sql <- "
  SELECT
    c.company,
    COUNT(DISTINCT sc.mmsi) AS n_ships,
    COUNT(*) AS n_records,
    SUM(sc.dist_travelled_km) AS total_km,
    SUM(sc.hours) AS total_hours
  FROM ship_cell sc
  JOIN ship s ON sc.mmsi = s.mmsi
  JOIN company c ON s.operator ILIKE '%' || c.operator_lookup || '%'
  GROUP BY c.company
  ORDER BY total_km DESC
"

company_summary <- DBI::dbGetQuery(con, company_sql)
company_summary |>
  mutate(
    pct_km    = total_km / sum(total_km) * 100,
    pct_hours = total_hours / sum(total_hours) * 100
  ) |>
  kable(
    digits = 1,
    format.args = list(big.mark = ","),
    col.names = c(
      "Company", "Ships", "Records", "Distance (km)", "Hours",
      "% Distance", "% Hours"
    )
  )
```

## Phase 2: Risk Calculations

### 2.1 Traffic by Speed Bin and Cell

```{r}
#| label: traffic-by-cell-speed
#| eval: false

# this query aggregates traffic by cell and speed bin
# for intersection with whale densities
traffic_sql <- glue("
  SELECT
    c.cell_id,
    {get_speed_bin_sql()} AS speed_bin,
    SUM(sc.dist_travelled_km) AS total_km,
    SUM(sc.hours) AS total_hours,
    COUNT(DISTINCT sc.mmsi) AS n_ships
  FROM cell c
  JOIN ship_cell sc ON c.cell_id = sc.cell_id
  WHERE sc.avg_speed_knots IS NOT NULL
  GROUP BY c.cell_id, speed_bin
")

# NOTE: execute this query and join with whale densities
# traffic_by_cell <- DBI::dbGetQuery(con, traffic_sql)
```

### 2.2 Risk by Area (placeholder)

```{r}
#| label: risk-by-area
#| eval: false

# TODO: implement spatial intersection and risk calculation
# 1. intersect whale hexagons with ship cell grid
# 2. intersect results with critical habitat and subareas
# 3. apply calc_speed_weighted_risk() function
# 4. summarize by area using summarize_risk_by_area()
```

## Deliverable Maps

### D1a: Rice's Whale Densities

```{r}
#| label: map-whale-density

# basic leaflet map of whale densities
leaflet() |>

  addProviderTiles("CartoDB.Positron") |>
  addPolygons(
    data = ply_goa,
    fillColor = "transparent",
    color = "gray",
    weight = 1,
    label = "Study Area"
  ) |>
  addPolygons(
    data = ply_crithab,
    fillColor = "blue",
    fillOpacity = 0.2,
    color = "blue",
    weight = 2,
    label = "Critical Habitat"
  ) |>
  addPolygons(
    data = ply_subareas,
    fillColor = "orange",
    fillOpacity = 0.3,
    color = "orange",
    weight = 2,
    label = ~as.character(1:nrow(ply_subareas))
  )
```

### D1b: Vessel Traffic (placeholder)

```{r}
#| label: map-traffic
#| eval: false

# TODO: create choropleth of vessel traffic by cell
```
### D1c: Strike Risk (placeholder)

```{r}
#| label: map-risk
#| eval: false

# TODO: create choropleth of strike risk by cell
```

## Deliverable Tables

### D2: Whale Densities by Area

```{r}
#| label: tbl-whale-density
#| eval: false

# TODO: summarize whale densities by study area and critical habitat
```

### D3: Risk by Area (Unweighted)

```{r}
#| label: tbl-risk-unweighted
#| eval: false

# TODO: base risk (whale x traffic, no speed weighting)
```

### D4: Risk by Area (Speed-Weighted)

```{r}
#| label: tbl-risk-weighted
#| eval: false

# TODO: Conn/Silber weighted risk
```

### D5: Risk Reduction Scenarios

```{r}
#| label: tbl-risk-reduction
#| eval: false

# TODO: risk reduction from slowdown scenarios
```

### D7: Company-Specific Results

```{r}
#| label: tbl-company-risk
#| eval: false

# TODO: per-company risk and reduction
```

## Cleanup

```{r}
#| label: cleanup

DBI::dbDisconnect(con, shutdown = TRUE)
```

## References

- Best (unpub.) - Previous ship-strike risk analysis for Rice's whale
- Benioff Ocean Science Laboratory (unpub.) - 2023 AIS analysis
- Conn, P.B. & Silber, G.K. (2013). Vessel speed restrictions reduce risk of collision-related mortality for North Atlantic right whales. *Ecosphere* 4(4):43.
- Litz, J.A., et al. (2022). Rice's whale (*Balaenoptera ricei*) spatial distribution, abundance, and status. *Marine Mammal Science* 38:1430-1451.
- NMFS (2020). Endangered Species Act biological opinion on the federally regulated oil and gas program activities in the Gulf of Mexico.
- NMFS (2023). Proposed rule to designate critical habitat for Rice's whale. 88 FR 47453.
