---
title: "Rice's Whale Ship-Strike Risk Analysis"
subtitle: "NRDC Contract - 2023 AIS Data Analysis"
author: "Ben Best, EcoQuants LLC"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    number-sections: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

```{r setup}
#| label: setup

# load libraries ----
librarian::shelf(
  DBI, dplyr, duckdb, glue, here, janitor, knitr, leaflet, leaflet.extras,
  lubridate, purrr, sf, stringr, terra, tibble, tidyr, viridisLite,
  quiet = TRUE
)

# source risk functions ----
source(here("scripts/risk_functions.R"))

# set paths ----
is_server   <- Sys.info()[["sysname"]] == "Linux"
dir_private <- ifelse(is_server, "/share/private", "~/My Drive/private")
dir_data    <- ifelse(is_server, "/share/data/ricei", "~/My Drive/projects/ricei/data")
dir_raw     <- glue("{dir_data}/raw")
goa_db      <- glue("{dir_data}/raw/ships/gulf_of_mexico_2023/goa.duckdb")
whales_shp  <- glue("{dir_raw}/whales/Rices_Whale_Monthly_Density.shp")

# connect to database ----
con <- DBI::dbConnect(duckdb::duckdb(), dbdir = goa_db, read_only = FALSE)
DBI::dbExecute(con, "INSTALL spatial; LOAD spatial;")
# dbListTables(con)  # "cell"       "company"    "goa"        "ship"       "ship_cell"  "whale"      "whale_cell"
```

## Overview

This analysis calculates ship-strike risk for Rice's whale (*Balaenoptera ricei*) in the Gulf of Mexico, using:

- **Whale densities:** Litz et al. (2022) hexagonal grid model
- **Vessel traffic:** 2023 AIS data from Benioff Ocean Science Laboratory
- **Risk model:** Conn & Silber (2013) speed-based risk factors

### Risk Factor Model

Based on Conn & Silber (2013), strike risk increases non-linearly with vessel speed:
```{r}
#| label: risk-factors-table

tibble(
  `Speed Bin`   = RISK_FACTORS$labels,
  `Risk Factor` = RISK_FACTORS$factors,
  `Rationale`   = c(
    "Baseline (slowdown target)",
    "Midpoint 11 kt",
    "Midpoint 13.5 kt",
    "Conservative at 15 kt"
  )
) |>
  kable()
```

## Data Inventory

### Database Tables

```{r}
#| label: db-inventory

# list tables
tables <- DBI::dbListTables(con)
cat("Tables in goa.duckdb:", paste(tables, collapse = ", "), "\n\n")

# count records
for (tbl_name in tables) {
  n <- tbl(con, tbl_name) |> count() |> collect() |> pull(n)
  cat(glue("{tbl_name}: {format(n, big.mark=',')} records"), "\n")
}
```

### Spatial Data Files

```{r}
#| label: spatial-inventory

dir_local_data <- here("data")

spatial_files <- c(
  "whales.geojson",
  "crithab_subareas.geojson",
  "nmfs_crit_hab.geojson",
  "goa.geojson"
)

for (f in spatial_files) {
  fpath <- file.path(dir_local_data, f)
  if (file.exists(fpath)) {
    sf_obj <- sf::st_read(fpath, quiet = TRUE)
    cat(glue("{f}: {nrow(sf_obj)} features, {ncol(sf_obj)} fields"), "\n")
  } else {
    cat(glue("{f}: NOT FOUND"), "\n")
  }
}
```

## Phase 1: Data Preparation

### 1.1 Load Spatial Boundaries

```{r}
#| label: load-spatial

# study area
ply_goa <- sf::st_read(here("data/goa.geojson"), quiet = TRUE)

# critical habitat
ply_crithab <- sf::st_read(here("data/nmfs_crit_hab.geojson"), quiet = TRUE)

# subareas (7 fairway/port approaches)
ply_subareas <- sf::st_read(here("data/crithab_subareas.geojson"), quiet = TRUE)

# whale densities
ply_whales <- sf::st_read(here("data/whales.geojson"), quiet = TRUE)

```

### 1.2 Import Whale Densities

Import the 2021 Rice's whale density data (Litz et al. 2022) into DuckDB tables using direct rasterization:

- `whale`: metadata for each month/metric combination
- `whale_cell`: cell-level density values matching the vessel grid

```{r}
#| label: import-whale-densities

# check if whale tables already exist
existing_tables <- DBI::dbListTables(con)
whale_exists      <- "whale" %in% existing_tables
whale_cell_exists <- "whale_cell" %in% existing_tables

if (!whale_exists || !whale_cell_exists) {
  cat("Whale tables not found. Importing using rasterization...\n")

  # create canonical raster template
  r_template <- create_cell_template(ply_goa)

  # process whale shapefile (rasterize to cell grid)
  whale_data <- process_whale_density(
    shp_path   = whales_shp,
    r_template = r_template,
    verbose    = TRUE
  )

  # import to database
  import_whale_to_duckdb(
    con            = con,
    whale_tbl      = whale_data$whale_tbl,
    whale_cell_tbl = whale_data$whale_cell_tbl,
    r_template     = r_template,
    overwrite      = TRUE,
    verbose        = TRUE
  )
} else {
  cat("Tables 'whale' and 'whale_cell' already exist in database.\n")
}

# create canonical raster template (if not already created above)
if (!exists("r_template"))
  r_template <- create_cell_template(ply_goa)

cat("\nRaster template:", nrow(r_template), "x", ncol(r_template), "cells\n")

# show summary if tables exist
if ("whale" %in% DBI::dbListTables(con)) {
  cat("\nWhale table summary:\n")
  DBI::dbGetQuery(con, "SELECT * FROM whale ORDER BY month_int, metric") |>
    kable(digits = 4)
}
```

### 1.3 Speed Distribution Summary

```{r}
#| label: speed-distribution

# query speed distribution with bins
speed_sql <- glue("
  SELECT
    {get_speed_bin_sql()} AS speed_bin,
    COUNT(*) AS n_records,
    SUM(dist_travelled_km) AS total_km,
    SUM(hours) AS total_hours
  FROM ship_cell
  WHERE avg_speed_knots IS NOT NULL
  GROUP BY speed_bin
  ORDER BY speed_bin
")

speed_dist <- DBI::dbGetQuery(con, speed_sql)
speed_dist |>
  mutate(
    pct_records = n_records / sum(n_records) * 100,
    pct_km      = total_km / sum(total_km) * 100,
    pct_hours   = total_hours / sum(total_hours) * 100
  ) |>
  kable(
    digits = 1,
    format.args = list(big.mark = ","),
    col.names = c(
      "Speed Bin", "Records", "Distance (km)", "Hours",
      "% Records", "% Distance", "% Hours"
    )
  )
```

### 1.3 Company Fleet Summary

```{r}
#| label: company-summary

# query company traffic in critical habitat
company_sql <- "
  SELECT
    c.company,
    COUNT(DISTINCT sc.mmsi) AS n_ships,
    COUNT(*) AS n_records,
    SUM(sc.dist_travelled_km) AS total_km,
    SUM(sc.hours) AS total_hours
  FROM ship_cell sc
  JOIN ship s ON sc.mmsi = s.mmsi
  JOIN company c ON s.operator ILIKE '%' || c.operator_lookup || '%'
  GROUP BY c.company
  ORDER BY total_km DESC
"

company_summary <- DBI::dbGetQuery(con, company_sql)
company_summary |>
  mutate(
    pct_km    = total_km / sum(total_km) * 100,
    pct_hours = total_hours / sum(total_hours) * 100
  ) |>
  kable(
    digits = 1,
    format.args = list(big.mark = ","),
    col.names = c(
      "Company", "Ships", "Records", "Distance (km)", "Hours",
      "% Distance", "% Hours"
    )
  )
```

### 1.4 Area Lookup (cell → geographic area)

Assign each raster cell to geographic areas (Study Area, Critical Habitat, Subareas 1–7) for grouped summaries.

```{r}
#| label: area-lookup

# rasterize subareas (cell_id → subarea_id)
r_sub <- terra::rasterize(
  terra::vect(ply_subareas), r_template, field = "subarea_id")
sub_df <- terra::as.data.frame(r_sub, cells = TRUE, na.rm = TRUE)
names(sub_df) <- c("cell_id", "subarea_id")

# rasterize critical habitat (cell_id → 1 if inside)
r_ch <- terra::rasterize(
  terra::vect(ply_crithab), r_template, field = 1)
ch_df <- terra::as.data.frame(r_ch, cells = TRUE, na.rm = TRUE)
names(ch_df) <- c("cell_id", "in_crithab")

# build lookup: every cell in DB is in the study area
area_lookup <- DBI::dbGetQuery(con, "SELECT DISTINCT cell_id FROM cell") |>
  left_join(sub_df, by = "cell_id") |>
  left_join(ch_df,  by = "cell_id") |>
  mutate(in_crithab = !is.na(in_crithab))

# expand to long format: one row per cell per area it belongs to
area_long <- bind_rows(
  # every cell → study area
  area_lookup |>
    mutate(area_label = "Study Area") |>
    select(cell_id, area_label),
  # critical habitat cells
  area_lookup |>
    filter(in_crithab) |>
    mutate(area_label = "Critical Habitat") |>
    select(cell_id, area_label),
  # subarea cells
  area_lookup |>
    filter(!is.na(subarea_id)) |>
    mutate(area_label = paste("Subarea", subarea_id)) |>
    select(cell_id, area_label)
)

cat("Area lookup cells:", nrow(area_lookup), "\n")
cat("Area-long rows:",    nrow(area_long), "\n")
area_long |> count(area_label) |> kable()
```

## Phase 2: Risk Calculations

### 2.1 Traffic by Speed Bin and Cell

```{r}
#| label: traffic-by-cell-speed

# aggregate traffic by cell and speed bin
traffic_sql <- glue("
  SELECT
    cell_id,
    {get_speed_bin_sql()} AS speed_bin,
    SUM(dist_travelled_km) AS total_km,
    SUM(hours) AS total_hours,
    COUNT(DISTINCT mmsi) AS n_ships
  FROM ship_cell
  WHERE avg_speed_knots IS NOT NULL
  GROUP BY cell_id, speed_bin
")

traffic_by_cell <- DBI::dbGetQuery(con, traffic_sql)
cat("Traffic rows:", nrow(traffic_by_cell),
    "| Cells:", n_distinct(traffic_by_cell$cell_id), "\n")
```

### 2.2 Risk by Area

```{r}
#| label: risk-by-area

# get annual whale density per cell
whale_annual <- DBI::dbGetQuery(con, "
  SELECT cell_id, value AS whale_density
  FROM whale_cell
  WHERE month_int = 0 AND metric = 'n'")

# join: traffic × whale density
risk_data <- traffic_by_cell |>
  inner_join(whale_annual, by = "cell_id") |>
  mutate(
    traffic   = total_km,
    speed_bin = factor(speed_bin, levels = RISK_FACTORS$labels))

cat("Risk data rows:", nrow(risk_data),
    "| Cells with both whale + traffic:", n_distinct(risk_data$cell_id), "\n")

# join with area long-format and summarize by area
risk_area_data <- risk_data |>
  inner_join(area_long, by = "cell_id", relationship = "many-to-many")

risk_by_area <- summarize_risk_by_area(
  risk_area_data,
  area_col = "area_label",
  verbose  = FALSE)

# order areas: study area first, then critical habitat, then subareas
area_order <- c("Study Area", "Critical Habitat",
                paste("Subarea", 1:7))
risk_by_area <- risk_by_area |>
  mutate(area_label = factor(area_label, levels = area_order)) |>
  arrange(area_label)

risk_by_area |> kable(digits = 2)
```

## Deliverable Maps

### D1a: Rice's Whale Densities (Annual Average)

Map of annual average whale density (n = number of individuals per cell) from Litz et al. (2022).

```{r}
#| label: map-whale-density-annual
#| fig-cap: "Annual average Rice's whale density from Litz et al. (2022) model. Values represent number of individuals per 0.1° cell."

# get annual average whale density raster
r_whale_annual <- get_whale_raster(
  con        = con,
  month_int  = 0,
  metric     = "n",
  r_template = r_template
)

# create map with critical habitat overlay
map_rast_simple(
  r             = r_whale_annual,
  legend_title  = "Whale Density<br><small>(n per cell)</small>",
  colors        = "Spectral",
  group         = "whale_density",
  overlay_sf    = ply_crithab,
  overlay_color = "blue",
  overlay_label = "Critical Habitat"
)
```

### D1a-monthly: Monthly Whale Densities

```{r}
#| label: map-whale-density-monthly
#| results: asis
#| fig-cap: "Monthly Rice's whale density maps showing seasonal variation."

# create monthly maps as tabs
cat("::: {.panel-tabset}\n\n")

for (m in 1:12) { # m = 1
  cat(glue("#### {month.abb[m]}\n\n"))

  r_month <- get_whale_raster(
    con        = con,
    month_int  = m,
    metric     = "n",
    r_template = r_template
  )

  map_out <- map_rast_simple(
    r            = r_month,
    legend_title = glue("Whale Density<br><small>({month.abb[m]})</small>"),
    colors       = "Spectral",
    group        = glue("whale_{month.abb[m]}")
  )

  # print map
  print(htmltools::tagList(map_out))
  cat("\n\n")
}

cat(":::\n")
```

### D1a-cv: Whale Density Uncertainty (CV)

Map of coefficient of variation (CV) for annual average whale density, showing spatial uncertainty in the model predictions.

```{r}
#| label: map-whale-cv
#| fig-cap: "Coefficient of variation (CV) for annual average whale density estimates."

# get annual average CV raster
r_whale_cv <- get_whale_raster(
  con        = con,
  month_int  = 0,
  metric     = "cv",
  r_template = r_template
)

map_rast_simple(
  r            = r_whale_cv,
  legend_title = "CV<br><small>(uncertainty)</small>",
  colors       = "YlOrRd",
  group        = "whale_cv"
)
```

### D1a-boundaries: Study Area Boundaries

```{r}
#| label: map-boundaries

# basic leaflet map of study area boundaries
leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addPolygons(
    data        = ply_goa,
    fillColor   = "transparent",
    color       = "gray",
    weight      = 1,
    label       = "Study Area"
  ) |>
  addPolygons(
    data        = ply_crithab,
    fillColor   = "blue",
    fillOpacity = 0.2,
    color       = "blue",
    weight      = 2,
    label       = "Critical Habitat"
  ) |>
  addPolygons(
    data        = ply_subareas,
    fillColor   = "orange",
    fillOpacity = 0.3,
    color       = "orange",
    weight      = 2,
    label       = ~as.character(1:nrow(ply_subareas))
  )
```

### D1b: Vessel Traffic

Map of total vessel traffic (distance traveled in km) per 0.1° cell from 2023 AIS data.

```{r}
#| label: map-traffic
#| fig-cap: "Vessel traffic density (total km traveled) from 2023 AIS data."

r_traffic <- get_traffic_raster(
  con        = con,
  metric     = "km",
  r_template = r_template)

map_rast_simple(
  r             = r_traffic,
  legend_title  = "Traffic<br><small>(km)</small>",
  colors        = "YlOrRd",
  group         = "traffic",
  overlay_sf    = ply_crithab,
  overlay_color = "blue",
  overlay_label = "Critical Habitat")
```
### D1c: Strike Risk

Map of speed-weighted ship-strike risk per cell (whale density × traffic × Conn/Silber risk factor).

```{r}
#| label: map-risk
#| fig-cap: "Speed-weighted ship-strike risk for Rice's whale. Risk = whale density × vessel km × Conn/Silber (2013) speed factor."

# aggregate risk to per-cell totals (speed-weighted)
risk_per_cell <- risk_data |>
  mutate(
    risk_factor = get_risk_factor(speed_bin),
    cell_risk   = whale_density * traffic * risk_factor) |>
  group_by(cell_id) |>
  summarize(risk = sum(cell_risk, na.rm = TRUE), .groups = "drop")

r_risk <- terra::init(r_template, NA)
r_risk[risk_per_cell$cell_id] <- risk_per_cell$risk
names(r_risk) <- "strike_risk"

map_rast_simple(
  r             = r_risk,
  legend_title  = "Strike Risk<br><small>(weighted)</small>",
  colors        = "YlOrRd",
  group         = "strike_risk",
  overlay_sf    = ply_crithab,
  overlay_color = "blue",
  overlay_label = "Critical Habitat")
```

## Deliverable Tables

### D2: Whale Densities by Area

```{r}
#| label: tbl-whale-density

# summarize whale densities by area
whale_by_area <- whale_annual |>
  inner_join(area_long, by = "cell_id", relationship = "many-to-many") |>
  group_by(area_label) |>
  summarize(
    n_cells       = n(),
    mean_density  = mean(whale_density, na.rm = TRUE),
    total_density = sum(whale_density, na.rm = TRUE),
    .groups       = "drop") |>
  mutate(area_label = factor(area_label,
    levels = c("Study Area", "Critical Habitat",
               paste("Subarea", 1:7)))) |>
  arrange(area_label)

whale_by_area |>
  kable(
    digits    = 4,
    col.names = c("Area", "Cells", "Mean Density", "Total Density"))
```

### D3: Risk by Area (Unweighted)

Base risk = whale density × traffic distance (no speed weighting).

```{r}
#| label: tbl-risk-unweighted

risk_by_area |>
  calc_area_contribution(risk_col = "total_base_risk") |>
  select(area_label, total_base_risk, contribution_pct) |>
  kable(
    digits    = c(0, 2, 1),
    col.names = c("Area", "Base Risk", "% of Total"),
    format.args = list(big.mark = ","))
```

### D4: Risk by Area (Speed-Weighted)

Speed-weighted risk using Conn & Silber (2013) factors.

```{r}
#| label: tbl-risk-weighted

risk_by_area |>
  calc_area_contribution(risk_col = "total_speed_weighted_risk") |>
  select(area_label, total_speed_weighted_risk, contribution_pct) |>
  kable(
    digits    = c(0, 2, 1),
    col.names = c("Area", "Speed-Weighted Risk", "% of Total"),
    format.args = list(big.mark = ","))
```

### D5: Risk Reduction Scenarios

Risk reduction if all vessels slowed to ≤10 kt (baseline Conn/Silber factor = 1.0).

```{r}
#| label: tbl-risk-reduction

risk_by_area |>
  select(area_label, total_speed_weighted_risk, total_slowdown_risk,
         risk_reduction_abs, risk_reduction_pct) |>
  kable(
    digits    = c(0, 2, 2, 2, 1),
    col.names = c("Area", "Current Risk", "Slowdown Risk",
                  "Reduction", "Reduction %"),
    format.args = list(big.mark = ","))
```

### D7: Company-Specific Results

Per-company risk and potential reduction from fleet-wide slowdown to ≤10 kt.

```{r}
#| label: tbl-company-risk

# query traffic by cell, speed bin, and company
company_traffic_sql <- glue("
  SELECT
    sc.cell_id,
    co.company,
    {get_speed_bin_sql()} AS speed_bin,
    SUM(sc.dist_travelled_km) AS total_km
  FROM ship_cell sc
  JOIN ship s ON sc.mmsi = s.mmsi
  JOIN company co ON s.operator ILIKE '%' || co.operator_lookup || '%'
  WHERE sc.avg_speed_knots IS NOT NULL
  GROUP BY sc.cell_id, co.company, speed_bin
")

company_traffic <- DBI::dbGetQuery(con, company_traffic_sql)

# join with whale density and compute risk
company_risk_data <- company_traffic |>
  inner_join(whale_annual, by = "cell_id") |>
  mutate(
    traffic   = total_km,
    speed_bin = factor(speed_bin, levels = RISK_FACTORS$labels))

# summarize by company
company_risk <- summarize_risk_by_area(
  company_risk_data,
  area_col = "company",
  verbose  = FALSE)

company_risk |>
  arrange(desc(total_speed_weighted_risk)) |>
  kable(
    digits    = c(0, 2, 2, 2, 2, 1),
    col.names = c("Company", "Base Risk", "Speed-Weighted Risk",
                  "Slowdown Risk", "Reduction", "Reduction %"),
    format.args = list(big.mark = ","))
```

## Cleanup

```{r}
#| label: cleanup

DBI::dbDisconnect(con, shutdown = TRUE)
```

## References

- Best (unpub.) - Previous ship-strike risk analysis for Rice's whale
- Benioff Ocean Science Laboratory (unpub.) - 2023 AIS analysis
- Conn, P.B. & Silber, G.K. (2013). Vessel speed restrictions reduce risk of collision-related mortality for North Atlantic right whales. *Ecosphere* 4(4):43.
- Litz, J.A., et al. (2022). Rice's whale (*Balaenoptera ricei*) spatial distribution, abundance, and status. *Marine Mammal Science* 38:1430-1451.
- NMFS (2020). Endangered Species Act biological opinion on the federally regulated oil and gas program activities in the Gulf of Mexico.
- NMFS (2023). Proposed rule to designate critical habitat for Rice's whale. 88 FR 47453.
