---
title: "Rice's Whale Ship-Strike Risk Analysis"
author: "Ben Best, EcoQuants LLC for NRDC"
date: today
bibliography: ricei.bib
nocite: |
  @nmfsBiologicalOpinionFederally2020, @nmfsProposedRuleCriticalHabitat2023
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    number-sections: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup

# load libraries ----
librarian::shelf(
  DBI, dplyr, duckdb, ggplot2, glue, here, janitor, knitr, leaflet,
  leaflet.extras, lubridate, purrr, scales, sf, stringr, terra, tibble,
  tidyr, viridisLite,
  quiet = TRUE
)

# source risk functions ----
source(here("scripts/risk_functions.R"))

# set paths ----
is_server   <- Sys.info()[["sysname"]] == "Linux"
dir_private <- ifelse(is_server, "/share/private", "~/My Drive/private")
dir_data    <- ifelse(is_server, "/share/data/ricei", "~/My Drive/projects/ricei/data")
dir_raw     <- glue("{dir_data}/raw")
goa_db      <- glue("{dir_data}/raw/ships/gulf_of_mexico_2023/goa.duckdb")
whales_shp  <- glue("{dir_raw}/whales/Rices_Whale_Monthly_Density.shp")

# connect to database ----
con <- DBI::dbConnect(duckdb::duckdb(), dbdir = goa_db, read_only = FALSE)
DBI::dbExecute(con, "INSTALL spatial; LOAD spatial;")
# dbListTables(con)  # "cell"       "company"    "goa"        "ship"       "ship_cell"  "whale"      "whale_cell"
```

## Overview

This analysis calculates ship-strike risk for Rice's whale (*Balaenoptera ricei*) in the Gulf of Mexico, using:

- **Whale densities:** @litzCetaceanSeaTurtle2022 hexagonal grid model
- **Vessel traffic:** 2023 AIS data from @benioffAISAnalysis2023
- **Risk model:** @connVesselSpeedRestrictions2013 speed-based risk factors

### Risk Factor Model

Based on @connVesselSpeedRestrictions2013, strike risk increases non-linearly with vessel speed:
```{r}
#| label: risk-factors-table

tibble(
  `Speed Bin`   = RISK_FACTORS$labels,
  `Risk Factor` = RISK_FACTORS$factors,
  `Rationale`   = c(
    "Baseline (slowdown target)",
    "Midpoint 11 kt",
    "Midpoint 13.5 kt",
    "Conservative at 15 kt"
  )
) |>
  kable()
```

## Data Inventory

### Database Tables

```{r}
#| label: db-inventory

# list tables
tables <- DBI::dbListTables(con)
cat("Tables in goa.duckdb:", paste(tables, collapse = ", "), "\n\n")

# count records
for (tbl_name in tables) {
  n <- tbl(con, tbl_name) |> count() |> collect() |> pull(n)
  cat(glue("{tbl_name}: {format(n, big.mark=',')} records"), "\n")
}
```

### Spatial Data Files

```{r}
#| label: spatial-inventory

dir_local_data <- here("data")

spatial_files <- c(
  "whales.geojson",
  "crithab_subareas.geojson",
  "nmfs_crit_hab.geojson",
  "goa.geojson"
)

for (f in spatial_files) {
  fpath <- file.path(dir_local_data, f)
  if (file.exists(fpath)) {
    sf_obj <- sf::st_read(fpath, quiet = TRUE)
    cat(glue("{f}: {nrow(sf_obj)} features, {ncol(sf_obj)} fields"), "\n")
  } else {
    cat(glue("{f}: NOT FOUND"), "\n")
  }
}
```

## Phase 1: Data Preparation

### 1.1 Load Spatial Boundaries

```{r}
#| label: load-spatial

# study area
ply_goa <- sf::st_read(here("data/goa.geojson"), quiet = TRUE)

# critical habitat
ply_crithab <- sf::st_read(here("data/nmfs_crit_hab.geojson"), quiet = TRUE)

# subareas (7 fairway/port approaches)
ply_subareas <- sf::st_read(here("data/crithab_subareas.geojson"), quiet = TRUE)

# whale densities
ply_whales <- sf::st_read(here("data/whales.geojson"), quiet = TRUE)

```

### 1.2 Import Whale Densities

Import the 2021 Rice's whale density data [@litzCetaceanSeaTurtle2022] into DuckDB tables using direct rasterization:

- `whale`: metadata for each month/metric combination
- `whale_cell`: cell-level density values matching the vessel grid

```{r}
#| label: import-whale-densities

# check if whale tables already exist
existing_tables <- DBI::dbListTables(con)
whale_exists      <- "whale" %in% existing_tables
whale_cell_exists <- "whale_cell" %in% existing_tables

if (!whale_exists || !whale_cell_exists) {
  cat("Whale tables not found. Importing using rasterization...\n")

  # create canonical raster template
  r_template <- create_cell_template(ply_goa)

  # process whale shapefile (rasterize to cell grid)
  whale_data <- process_whale_density(
    shp_path   = whales_shp,
    r_template = r_template,
    verbose    = TRUE
  )

  # import to database
  import_whale_to_duckdb(
    con            = con,
    whale_tbl      = whale_data$whale_tbl,
    whale_cell_tbl = whale_data$whale_cell_tbl,
    r_template     = r_template,
    overwrite      = TRUE,
    verbose        = TRUE
  )
} else {
  cat("Tables 'whale' and 'whale_cell' already exist in database.\n")
}

# create canonical raster template (if not already created above)
if (!exists("r_template"))
  r_template <- create_cell_template(ply_goa)

cat("\nRaster template:", nrow(r_template), "x", ncol(r_template), "cells\n")

# show summary if tables exist
if ("whale" %in% DBI::dbListTables(con)) {
  cat("\nWhale table summary:\n")
  DBI::dbGetQuery(con, "SELECT * FROM whale ORDER BY month_int, metric") |>
    kable(digits = 4)
}
```

### 1.3 Speed Distribution Summary

```{r}
#| label: speed-distribution

# query speed distribution with bins
speed_sql <- glue("
  SELECT
    {get_speed_bin_sql()} AS speed_bin,
    COUNT(*) AS n_records,
    SUM(dist_travelled_km) AS total_km,
    SUM(hours) AS total_hours
  FROM ship_cell
  WHERE avg_speed_knots IS NOT NULL
  GROUP BY speed_bin
  ORDER BY speed_bin
")

speed_dist <- DBI::dbGetQuery(con, speed_sql)
speed_dist |>
  mutate(
    pct_records = n_records / sum(n_records) * 100,
    pct_km      = total_km / sum(total_km) * 100,
    pct_hours   = total_hours / sum(total_hours) * 100
  ) |>
  kable(
    digits = 1,
    format.args = list(big.mark = ","),
    col.names = c(
      "Speed Bin", "Records", "Distance (km)", "Hours",
      "% Records", "% Distance", "% Hours"
    )
  )
```

### 1.3 Company Fleet Summary

```{r}
#| label: company-summary

# query company traffic in critical habitat
company_sql <- "
  SELECT
    c.company,
    COUNT(DISTINCT sc.mmsi) AS n_ships,
    COUNT(*) AS n_records,
    SUM(sc.dist_travelled_km) AS total_km,
    SUM(sc.hours) AS total_hours
  FROM ship_cell sc
  JOIN ship s ON sc.mmsi = s.mmsi
  JOIN company c ON s.operator ILIKE '%' || c.operator_lookup || '%'
  GROUP BY c.company
  ORDER BY total_km DESC
"

company_summary <- DBI::dbGetQuery(con, company_sql)
company_summary |>
  mutate(
    pct_km    = total_km / sum(total_km) * 100,
    pct_hours = total_hours / sum(total_hours) * 100
  ) |>
  kable(
    digits = 1,
    format.args = list(big.mark = ","),
    col.names = c(
      "Company", "Ships", "Records", "Distance (km)", "Hours",
      "% Distance", "% Hours"
    )
  )
```

### 1.4 Area Lookup (cell → geographic area)

Assign each raster cell to geographic areas (Study Area, Critical Habitat, Subareas 1–7) for grouped summaries.

```{r}
#| label: area-lookup

# rasterize subareas (cell_id → subarea_id)
r_sub <- terra::rasterize(
  terra::vect(ply_subareas), r_template, field = "subarea_id")
sub_df <- terra::as.data.frame(r_sub, cells = TRUE, na.rm = TRUE)
names(sub_df) <- c("cell_id", "subarea_id")

# rasterize critical habitat (cell_id → 1 if inside)
r_ch <- terra::rasterize(
  terra::vect(ply_crithab), r_template, field = 1)
ch_df <- terra::as.data.frame(r_ch, cells = TRUE, na.rm = TRUE)
names(ch_df) <- c("cell_id", "in_crithab")

# build lookup: every cell in DB is in the study area
area_lookup <- DBI::dbGetQuery(con, "SELECT DISTINCT cell_id FROM cell") |>
  left_join(sub_df, by = "cell_id") |>
  left_join(ch_df,  by = "cell_id") |>
  mutate(in_crithab = !is.na(in_crithab))

# expand to long format: one row per cell per area it belongs to
area_long <- bind_rows(
  # every cell → study area
  area_lookup |>
    mutate(area_label = "Study Area") |>
    select(cell_id, area_label),
  # critical habitat cells
  area_lookup |>
    filter(in_crithab) |>
    mutate(area_label = "Critical Habitat") |>
    select(cell_id, area_label),
  # subarea cells
  area_lookup |>
    filter(!is.na(subarea_id)) |>
    mutate(area_label = paste("Subarea", subarea_id)) |>
    select(cell_id, area_label)
)

cat("Area lookup cells:", nrow(area_lookup), "\n")
cat("Area-long rows:",    nrow(area_long), "\n")
area_long |> count(area_label) |> kable()
```

## Phase 2: Risk Calculations

### 2.1 Traffic by Speed Bin and Cell

```{r}
#| label: traffic-by-cell-speed

# aggregate traffic by cell and speed bin
traffic_sql <- glue("
  SELECT
    cell_id,
    {get_speed_bin_sql()} AS speed_bin,
    SUM(dist_travelled_km) AS total_km,
    SUM(hours) AS total_hours,
    COUNT(DISTINCT mmsi) AS n_ships
  FROM ship_cell
  WHERE avg_speed_knots IS NOT NULL
  GROUP BY cell_id, speed_bin
")

traffic_by_cell <- DBI::dbGetQuery(con, traffic_sql)
cat("Traffic rows:", nrow(traffic_by_cell),
    "| Cells:", n_distinct(traffic_by_cell$cell_id), "\n")
```

### 2.2 Risk by Area

```{r}
#| label: risk-by-area

# get annual whale density per cell
whale_annual <- DBI::dbGetQuery(con, "
  SELECT cell_id, value AS whale_density
  FROM whale_cell
  WHERE month_int = 0 AND metric = 'n'")

# join: traffic × whale density
risk_data <- traffic_by_cell |>
  inner_join(whale_annual, by = "cell_id") |>
  mutate(
    traffic   = total_km,
    speed_bin = factor(speed_bin, levels = RISK_FACTORS$labels))

cat("Risk data rows:", nrow(risk_data),
    "| Cells with both whale + traffic:", n_distinct(risk_data$cell_id), "\n")

# join with area long-format and summarize by area
risk_area_data <- risk_data |>
  inner_join(area_long, by = "cell_id", relationship = "many-to-many")

risk_by_area <- summarize_risk_by_area(
  risk_area_data,
  area_col = "area_label",
  verbose  = FALSE)

# order areas: study area first, then critical habitat, then subareas
area_order <- c("Study Area", "Critical Habitat",
                paste("Subarea", 1:7))
risk_by_area <- risk_by_area |>
  mutate(area_label = factor(area_label, levels = area_order)) |>
  arrange(area_label)

risk_by_area |> kable(digits = 2)
```

## Phase 3: Garrison et al. (2025) Lethality-Curve Risk

This section adds a continuous P(lethality | speed, vessel_size) approach from @garrisonLethalityCurveGLM2025, as implemented in [SEFSC/VesselStrikeRiskModel](https://github.com/SEFSC/VesselStrikeRiskModel). Unlike the @connVesselSpeedRestrictions2013 discrete speed bins, this GLM produces a smooth lethality curve that also accounts for vessel size class.

**Simplified formula:**

$$\text{Risk}_\text{Garrison} = \text{whale\_density} \times \text{dist\_km} \times P(\text{lethality} \mid \text{speed}, \text{vessel\_size})$$

::: {.callout-note collapse="true"}
## Caveats

The full @blondinVesselStrikeMortality2025 encounter-to-mortality model computes:

$$\text{Mortality} = \text{EncounterRate} \times P(\text{surface}) \times P(\text{strike}) \times P(\text{lethality}) \times (1 - P(\text{avoidance}))$$

We implement **only P(lethality)** because we lack:

- **Encounter rate model** — requires individual vessel track geometry (we have cell-aggregated transits)
- **P(surface)** — requires Rice's whale dive profile data [only 1 tagged whale; @kokDiveBehaviorRicesWhale2023]
- **P(avoidance)** — requires species-specific reaction parameters not yet available for Rice's whale

Additional notes:

- The Garrison GLM was fit on NARW + other large whale strike data, not Rice's whale specifically
- Rice's whale is modeled as `spe.HB = "Not Humpback"` (appropriate given it is not a humpback)
- Vessels with missing `length_m` default to class "L" (most common for AIS-tracked commercial vessels)
- We use `avg_speed_knots` per cell transit, not instantaneous speed

See **Phase 4** for a comprehensive evaluation of all Blondin model parameters and a diel risk scenario analysis using the single available Rice's whale dive profile.
:::

### 3.0 Load Garrison GLM

```{r}
#| label: garrison-load-glm

garrison_glm <- get_garrison_glm()

cat("Garrison lethality GLM summary:\n")
cat("  Family:", family(garrison_glm)$family, "\n")
cat("  Link:  ", family(garrison_glm)$link, "\n")
cat("  AIC:   ", AIC(garrison_glm), "\n\n")
cat("Coefficients:\n")
print(coef(garrison_glm))
```

### 3.1 Vessel Size Distribution

```{r}
#| label: garrison-vessel-sizes

# deduplicate ships by mmsi (most recent record)
vessel_size_sql <- glue("
  WITH ship_dedup AS (
    SELECT *, ROW_NUMBER() OVER (
      PARTITION BY mmsi ORDER BY updated_date DESC
    ) AS rn
    FROM ship
  )
  SELECT
    (s.length_m IS NOT NULL) AS has_length,
    {get_vessel_size_sql()} AS vessel_size,
    COUNT(DISTINCT sc.mmsi) AS n_ships,
    SUM(sc.dist_travelled_km) AS total_km
  FROM ship_cell sc
  LEFT JOIN ship_dedup s ON sc.mmsi = s.mmsi AND s.rn = 1
  WHERE sc.avg_speed_knots IS NOT NULL
  GROUP BY has_length, vessel_size
  ORDER BY vessel_size
", .open = "{", .close = "}")

# note: LEFT JOIN keeps ship_cell rows even without ship metadata
vessel_sizes <- DBI::dbGetQuery(con, vessel_size_sql)

# summarize by vessel size class
vessel_size_summary <- vessel_sizes |>
  group_by(vessel_size) |>
  summarize(
    n_ships  = sum(n_ships),
    total_km = sum(total_km),
    .groups  = "drop") |>
  mutate(
    pct_ships = n_ships / sum(n_ships) * 100,
    pct_km    = total_km / sum(total_km) * 100)

# report coverage: vessels with known length_m
total_km <- sum(vessel_sizes$total_km)
known_km <- vessel_sizes |>
  filter(has_length) |>
  pull(total_km) |> sum()
cat(glue(
  "Vessel length coverage: {round(known_km/total_km*100, 1)}% of ",
  "ship_cell traffic km from vessels with known length_m\n\n"))

vessel_size_summary |>
  kable(
    digits    = c(0, 0, 0, 1, 1),
    col.names = c("Vessel Size", "Ships", "Total km", "% Ships", "% km"),
    format.args = list(big.mark = ","))
```

### 3.2 Lethality Curve Plot

```{r}
#| label: garrison-lethality-curve
#| fig-cap: "@garrisonLethalityCurveGLM2025 P(lethality) as a function of vessel speed, by vessel size class. Vertical dashed line marks the 10-kt slowdown target."

# generate prediction grid: speeds 0–25 kt × 4 size classes
speed_grid <- expand.grid(
  speed       = seq(0, 25, by = 0.5),
  vessel_size = c("S", "M", "L", "XL"))

speed_grid$p_lethal <- calc_p_lethal_garrison(
  speed_knots = speed_grid$speed,
  vessel_size = speed_grid$vessel_size,
  glm_model   = garrison_glm)

# order vessel_size factor for legend
speed_grid$vessel_size <- factor(
  speed_grid$vessel_size,
  levels = c("S", "M", "L", "XL"))

ggplot2::ggplot(speed_grid, ggplot2::aes(
    x = speed, y = p_lethal, color = vessel_size)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::geom_vline(xintercept = 10, linetype = "dashed", color = "gray40") +
  ggplot2::annotate("text", x = 10.5, y = 0.05, label = "10 kt slowdown",
                    hjust = 0, color = "gray40", size = 3) +
  ggplot2::scale_color_manual(
    values = c(S = "#2166ac", M = "#67a9cf", L = "#ef8a62", XL = "#b2182b"),
    name   = "Vessel Size") +
  ggplot2::scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  ggplot2::labs(
    x = "Vessel Speed (knots)",
    y = "P(lethality)") +
  ggplot2::theme_minimal()
```

### 3.3 Garrison Traffic Query

```{r}
#| label: garrison-traffic-query

# query per-record traffic with vessel size classification
garrison_traffic_sql <- glue("
  WITH ship_dedup AS (
    SELECT *, ROW_NUMBER() OVER (
      PARTITION BY mmsi ORDER BY updated_date DESC
    ) AS rn
    FROM ship
  )
  SELECT
    sc.cell_id,
    sc.mmsi,
    sc.avg_speed_knots,
    sc.dist_travelled_km,
    {get_vessel_size_sql()} AS vessel_size
  FROM ship_cell sc
  LEFT JOIN ship_dedup s ON sc.mmsi = s.mmsi AND s.rn = 1
  WHERE sc.avg_speed_knots IS NOT NULL
", .open = "{", .close = "}")

garrison_traffic <- DBI::dbGetQuery(con, garrison_traffic_sql)

cat("Garrison traffic records:", format(nrow(garrison_traffic), big.mark = ","),
    "| Cells:", n_distinct(garrison_traffic$cell_id),
    "| Vessels:", n_distinct(garrison_traffic$mmsi), "\n")
```

### 3.4 Compute P(lethality) and Risk

```{r}
#| label: garrison-compute-risk

# compute P(lethality) at current speed and at 10-kt slowdown
garrison_traffic <- garrison_traffic |>
  mutate(
    p_lethal      = calc_p_lethal_garrison(
      avg_speed_knots, vessel_size, garrison_glm),
    speed_slow    = pmin(avg_speed_knots, 10),
    p_lethal_slow = calc_p_lethal_garrison(
      speed_slow, vessel_size, garrison_glm) )

# join with whale density
garrison_risk <- garrison_traffic |>
  inner_join(whale_annual, by = "cell_id") |>
  mutate(
    traffic              = dist_travelled_km,
    garrison_risk        = whale_density * dist_travelled_km * p_lethal,
    garrison_slowdown_risk = whale_density * dist_travelled_km * p_lethal_slow)

cat("Records with whale overlap:", format(nrow(garrison_risk), big.mark = ","),
    "| Cells:", n_distinct(garrison_risk$cell_id), "\n")
cat("Total Garrison risk:  ", format(sum(garrison_risk$garrison_risk), big.mark = ","), "\n")
cat("Total slowdown risk:  ", format(sum(garrison_risk$garrison_slowdown_risk), big.mark = ","), "\n")
```

### 3.5 Risk by Area (Garrison)

```{r}
#| label: garrison-risk-by-area

# join with area lookup and summarize
garrison_area_data <- garrison_risk |>
  inner_join(area_long, by = "cell_id", relationship = "many-to-many")

garrison_by_area <- summarize_garrison_risk(
  garrison_area_data,
  area_col = "area_label",
  verbose  = FALSE)

# order areas
garrison_by_area <- garrison_by_area |>
  mutate(area_label = factor(area_label, levels = area_order)) |>
  arrange(area_label)

garrison_by_area |>
  kable(
    digits      = c(0, 2, 4, 4, 4, 1),
    col.names   = c("Area", "Base Risk", "Garrison Risk",
                     "Slowdown Risk", "Reduction", "Reduction %"),
    format.args = list(big.mark = ","))
```

### 3.6 Garrison Risk Map (D1c-g)

```{r}
#| label: map-garrison-risk
#| fig-cap: "@garrisonLethalityCurveGLM2025 lethality-weighted ship-strike risk. Risk = whale density × vessel km × P(lethality | speed, vessel size)."

# aggregate to per-cell totals
garrison_per_cell <- garrison_risk |>
  group_by(cell_id) |>
  summarize(risk = sum(garrison_risk, na.rm = TRUE), .groups = "drop")

r_garrison_risk <- terra::init(r_template, NA)
r_garrison_risk[garrison_per_cell$cell_id] <- garrison_per_cell$risk
names(r_garrison_risk) <- "garrison_risk"

map_rast_simple(
  r             = r_garrison_risk,
  legend_title  = "Garrison Risk<br><small>(P(lethal) weighted)</small>",
  colors        = "YlOrRd",
  group         = "garrison_risk",
  overlay_sf    = ply_crithab,
  overlay_color = "blue",
  overlay_label = "Critical Habitat")
```

### 3.7 Garrison Risk Reduction (D5-g)

```{r}
#| label: tbl-garrison-reduction

garrison_by_area |>
  select(area_label, total_garrison_risk, total_slowdown_risk,
         risk_reduction_abs, risk_reduction_pct) |>
  kable(
    digits      = c(0, 4, 4, 4, 1),
    col.names   = c("Area", "Current Risk", "Slowdown Risk",
                     "Reduction", "Reduction %"),
    format.args = list(big.mark = ","))
```

### 3.8 Company Garrison Risk (D7-g)

```{r}
#| label: garrison-company-risk

# query per-record traffic with vessel size + company
garrison_company_sql <- glue("
  WITH ship_dedup AS (
    SELECT *, ROW_NUMBER() OVER (
      PARTITION BY mmsi ORDER BY updated_date DESC
    ) AS rn
    FROM ship
  )
  SELECT
    sc.cell_id,
    sc.mmsi,
    co.company,
    sc.avg_speed_knots,
    sc.dist_travelled_km,
    {get_vessel_size_sql()} AS vessel_size
  FROM ship_cell sc
  LEFT JOIN ship_dedup s ON sc.mmsi = s.mmsi AND s.rn = 1
  JOIN company co ON s.operator ILIKE '%' || co.operator_lookup || '%'
  WHERE sc.avg_speed_knots IS NOT NULL
", .open = "{", .close = "}")

garrison_company <- DBI::dbGetQuery(con, garrison_company_sql)

# compute P(lethality) and risk
garrison_company <- garrison_company |>
  mutate(
    p_lethal      = calc_p_lethal_garrison(
      avg_speed_knots, vessel_size, garrison_glm),
    speed_slow    = pmin(avg_speed_knots, 10),
    p_lethal_slow = calc_p_lethal_garrison(
      speed_slow, vessel_size, garrison_glm) )

garrison_company_risk <- garrison_company |>
  inner_join(whale_annual, by = "cell_id") |>
  mutate(
    traffic                = dist_travelled_km,
    garrison_risk          = whale_density * dist_travelled_km * p_lethal,
    garrison_slowdown_risk = whale_density * dist_travelled_km * p_lethal_slow)

# summarize by company
company_garrison <- summarize_garrison_risk(
  garrison_company_risk,
  area_col = "company",
  verbose  = FALSE)

company_garrison |>
  arrange(desc(total_garrison_risk)) |>
  kable(
    digits      = c(0, 2, 4, 4, 4, 1),
    col.names   = c("Company", "Base Risk", "Garrison Risk",
                     "Slowdown Risk", "Reduction", "Reduction %"),
    format.args = list(big.mark = ","))
```

### 3.9 Comparison: Conn/Silber vs Garrison

```{r}
#| label: garrison-comparison

# build side-by-side comparison
comparison <- risk_by_area |>
  select(
    area_label,
    cs_risk     = total_speed_weighted_risk,
    cs_slowdown = total_slowdown_risk,
    cs_red_pct  = risk_reduction_pct) |>
  inner_join(
    garrison_by_area |>
      select(
        area_label,
        g_risk     = total_garrison_risk,
        g_slowdown = total_slowdown_risk,
        g_red_pct  = risk_reduction_pct),
    by = "area_label")

comparison |>
  kable(
    digits    = c(0, 2, 2, 1, 4, 4, 1),
    col.names = c(
      "Area",
      "C/S Risk", "C/S Slowdown", "C/S Reduction %",
      "Garrison Risk", "Garrison Slowdown", "Garrison Reduction %"),
    format.args = list(big.mark = ","))
```

The @connVesselSpeedRestrictions2013 approach uses discrete speed-bin multipliers (1.0–10.8×), while the @garrisonLethalityCurveGLM2025 GLM produces a continuous lethality probability that also depends on vessel size. Key differences:

- **Garrison risk reduction is generally smaller** for areas dominated by large/XL vessels, because P(lethality) for large vessels remains elevated even at low speeds
- **The two methods agree directionally**: slowdown to 10 kt reduces risk in all areas under both models
- **Garrison provides finer resolution**: vessels of different sizes in the same speed bin receive different risk weights

## Phase 4: Blondin et al. (2025) Full Model Parameter Evaluation

The full @blondinVesselStrikeMortality2025 encounter-to-mortality model estimates annual vessel-strike mortality as:

$$M_\text{year} = \sum_{i,j,k} \lambda_{ijk} \times P(\text{strike\_depth}) \times (1 - P(\text{avoid})) \times P(\text{lethality}) \times N_w$$

where $\lambda_{ijk}$ is the encounter rate for vessel $i$ in cell $j$ during month $k$, $P(\text{strike\_depth})$ is the probability the whale is within the vessel's strike zone, $P(\text{avoid})$ is the probability the whale avoids the strike through vertical descent, $P(\text{lethality})$ is the probability the strike is lethal (from the Garrison GLM), and $N_w$ is whale abundance.

Our Phase 3 analysis implements **only P(lethality)**, multiplied by whale density and vessel distance as a co-occurrence proxy. This section evaluates all parameters in the Blondin framework to document what is feasible for Rice's whale and what data gaps remain.

### 4.1 Parameter Evaluation Summary

```{r}
#| label: blondin-parameter-table

param_eval <- tibble::tibble(
  Parameter = c(
    "P(lethality | speed, vessel_size)",
    "Whale density (N_w)",
    "Vessel traffic (AIS)",
    "Vessel size classification",
    "Vessel draft & strike zone",
    "P(whale at strike depth)",
    "Encounter rate (λ)",
    "P(avoidance)",
    "Bathymetry (bottom depth)",
    "Whale height",
    "Small vessel AIS correction",
    "Monthly temporal resolution",
    "Diel (day/night) risk scenarios"
  ),
  Status = c(
    "\u2705 Implemented",
    "\u2705 Implemented",
    "\u2705 Implemented",
    "\u2705 Implemented",
    "\u26a0\ufe0f Feasible to estimate",
    "\u26a0\ufe0f Very limited data",
    "\u274c Not feasible",
    "\u274c No Rice's whale data",
    "\u2705 Data available",
    "\u26a0\ufe0f Assumed from related spp.",
    "\u274c No GoM registry data",
    "\u2705 Available, not yet used",
    "\u26a0\ufe0f Feasible as scenario analysis"
  ),
  `In Analysis?` = c(
    "Yes (Phase 3)",
    "Yes (Litz 2022)",
    "Yes (2023 AIS)",
    "Yes (S/M/L/XL)",
    "No",
    "No (assumes 1.0)",
    "No",
    "No (assumes 0.0)",
    "No",
    "No",
    "No",
    "No (annual avg)",
    "Yes (below)"
  ),
  Rationale = c(
    "Garrison GLM; species = 'Not Humpback' appropriate for Rice's whale",
    "Litz et al. (2022) monthly hex grid; newer Garrison et al. (2024) SDM also exists",
    "2023 Benioff AIS; cell-aggregated dist_km, hours, speed per cell per day",
    "~90% of vessel traffic km have known length_m; S/M/L/XL per Garrison thresholds",
    "Draft derivable from size class using Garrison defaults (S/M/L: 5 m, XL: 15 m)",
    "2 tagged Rice's whales (Kok et al. 2023); 7 more tagged in 2024 (data pending)",
    "Requires individual vessel track geometry; current data is cell-aggregated",
    "McKenna et al. (2015) blue whale parameters not transferable to Rice's whale",
    "GEBCO 2023 bathymetry available; critical habitat is 100-400 m depth",
    "Blondin uses 3.0 m for NARW (Moore et al. 2005); reasonable for Rice's whale",
    "Correction ~3.5x for vessels <65 ft; most critical habitat traffic is large commercial",
    "Monthly whale densities and ship dates in database; higher computational cost",
    "Kok et al. (2023) shows strong diel vertical migration; high management relevance"
  )
)

kable(param_eval)
```

### 4.2 Parameter Details

#### a. P(lethality | speed, vessel_size) — Implemented

The @garrisonLethalityCurveGLM2025 GLM (implemented in Phase 3) estimates the probability that a vessel strike is lethal as a function of vessel speed and size class. The model was fit on North Atlantic right whale (NARW) and other large whale strike data — not Rice's whale specifically. Rice's whale is modeled as `spe.HB = "Not Humpback"`, which is appropriate given it is a balaenopterid, not a humpback.

**Caveat:** The lethality curve may not perfectly represent Rice's whale biomechanics, but no species-specific data exist to fit an alternative.

#### b. Whale density (N_w) — Implemented

The @litzCetaceanSeaTurtle2022 monthly hexagonal grid model provides the best available spatial abundance estimates for Rice's whale. Our analysis uses the annual average (month_int = 0), though monthly values are also in the database.

**Caveat:** The model is based on 2021 survey data. @garrisonUpdatedSpeciesDistribution2024 have since developed an updated species distribution model that could be incorporated in future analyses.

#### c. Vessel traffic (AIS) — Implemented

We use 2023 AIS data from the Benioff Ocean Science Laboratory, aggregated to 0.1° cells with daily sums of distance traveled (km), hours, and average speed per MMSI per cell.

**Caveat:** Cell-aggregated data does not capture individual vessel transit geometry, which the full Blondin encounter model requires.

#### d. Vessel size classification — Implemented

Vessels are classified as S/M/L/XL based on `ship.length_m` using the @garrisonLethalityCurveGLM2025 thresholds. Approximately 90% of vessel traffic (by km) comes from vessels with known length. Vessels with missing length default to class "L".

#### e. Vessel draft and strike zone — Feasible to estimate

Vessel draft is not directly recorded in our database but the SEFSC `Step1_RiskModel.R` (lines 86–91) defines depth limits (`depth.lim`) by vessel size class:

- **M/L vessels:** `depth.lim = 5` m
- **XL vessels:** `depth.lim = 15` m (hull draft; `PropSuctDepth` is commented out in the code)

This means XL vessels have a strike zone 3× deeper than M/L vessels, making vessel size class analytically important. The code also includes a commented-out propeller suction depth (`PropSuctDepth = 1 × mean.draft`) that would extend the XL danger zone to ~30 m, but this is not active in the current model.

**Caveat:** Actual draft varies by vessel loading and type; these are average defaults.

#### f. P(whale at strike depth) — Very limited data

This parameter requires tag-derived dive profiles showing time-at-depth. For NARW, @blondinVesselStrikeMortality2025 used data from 146 depth-recording tags across 4 regions. For Rice's whale, @kokDiveBehaviorRicesWhale2023 presents dive profiles from two tagged individuals — Milky Way (2015) and Edna (2018) — both showing strong diel vertical migration (shallow at night, deep 200–270 m by day).

NMFS tagged 7 additional Rice's whales during a 2024 research cruise, but those data are not yet publicly available. Our current analysis implicitly assumes P(surface) = 1.0, which overstates risk.

**Caveat:** n = 2 whales; any population-level estimate carries substantial uncertainty. The diel scenario analysis (Section 4.3) uses digitized values from both whales to explore sensitivity to this parameter.

#### g. Encounter rate (λ) — Not feasible with current data

The Blondin 2D encounter rate model requires individual vessel transit time through each cell (track-level geometry). Our data format provides cell-aggregated daily sums per MMSI per cell, which lacks the spatial resolution for true encounter probability calculations.

**Rationale:** Using `whale_density × dist_km` as a co-occurrence proxy is reasonable for relative risk comparison across areas, even though it does not capture true encounter mechanics.

#### h. P(avoidance) — No Rice's whale behavioral data

The avoidance probability is based on @mckennaSimultaneousTrackingBlue2015 blue whale vertical avoidance behavior, requiring parameters for whale descent rate, reaction distance, and starting depth — none of which are available for Rice's whale.

Blondin's NARW values range from 0.81–2.0 m/s descent rate and 10–1200 m reaction distance. Their sensitivity analysis showed that **P(avoidance) is the largest single source of model uncertainty**: assuming zero avoidance vs. full avoidance changes mortality estimates by orders of magnitude.

**Caveat:** Our analysis implicitly assumes P(avoidance) = 0 (no avoidance), which provides an upper-bound risk estimate. Using NARW behavioral parameters for Rice's whale would introduce unknown directional bias.

#### i. Bathymetry (bottom depth) — Data available

GEBCO 2023 bathymetry data exists in the project (`data/raw/depth/gebco_2023_*.tif`). In shallow waters where bottom depth < strike zone + whale height, vertical avoidance is physically impossible. For XL vessels (strike zone = 15 m + 3 m whale height = 18 m), areas shallower than 18 m represent "unavoidable strike zones."

**Caveat:** Most Rice's whale critical habitat is 100–400 m deep, so shallow-water unavoidable zones are less relevant than for coastal NARW populations.

#### j. Whale height — Assumed from related species

@blondinVesselStrikeMortality2025 uses 3.0 m for NARW based on @mooreMorphometryGrossMorphology2005. Rice's whale is a similar-sized baleen whale, so 3 m is a reasonable assumption. This parameter is used in the avoidance calculation: a whale needs the strike zone depth plus its own body height of water column to descend clear of the vessel.

#### k. Small vessel AIS correction factor — No GoM registry data

@blondinVesselStrikeMortality2025 apply an approximately 3.5× correction factor for vessels < 65 ft that are not required to carry AIS transponders. Computing this correction for the Gulf of Mexico would require regional vessel registration data.

**Rationale:** Most traffic through Rice's whale critical habitat is from large commercial vessels that carry AIS, so the undercount primarily affects recreational and small fishing vessel risk estimates.

#### l. Monthly temporal resolution — Data available, not yet used

Both monthly whale densities (month_int 1–12) and vessel traffic dates are available in the database. The current analysis uses the annual average, but monthly analysis could reveal temporal risk hotspots where seasonal whale concentrations coincide with high traffic periods.

**Caveat:** Monthly analysis increases computational cost approximately 12×.

#### m. Diel (day/night) risk scenarios — Feasible as scenario analysis

@kokDiveBehaviorRicesWhale2023 demonstrated that the tagged Rice's whale ("Milky Way") spent significantly more time near the surface at night compared to daytime, consistent with diel vertical migration for foraging. Although time-of-day is not available for individual AIS records, we can compute risk under bounding scenarios to quantify the **diel risk differential** — how much more dangerous nighttime transits are compared to daytime. This approach was also referenced in the revised NRDC statement of work (Jasny, Mar 2025).

### 4.3 Diel Risk Scenario Analysis

Using dive profiles from @kokDiveBehaviorRicesWhale2023 for two tagged Rice's whales — Milky Way (2015) and Edna (2018) — we digitized the cumulative % time at depth from Fig. 5 using [WebPlotDigitizer](https://automeris.io/WebPlotDigitizer). These values represent P(surface): the proportion of time each whale spent at or above a given depth threshold during day vs. night.

```{r}
#| label: diel-p-surface

# get P(surface) estimates digitized from Kok et al. (2023) Fig. 5
p_surface <- get_p_surface_riceswhale(verbose = TRUE)

p_surface |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  kable(
    col.names = c(
      "Depth (m)", "Vessel Sizes",
      "M.Way Day", "M.Way Night",
      "Edna Day", "Edna Night",
      "Avg Day", "Avg Night", "Method"),
    caption = "P(surface): cumulative proportion of time at or above depth, from Kok et al. (2023) Fig. 5. Milky Way (2015) and Edna (2018). The 30 m row is extrapolated from the 15-to-20 m per-meter rate.")
```

For the scenario analysis, we use the **average across both whales** and apply P(surface) values matched to vessel size class per the SEFSC `Step1_RiskModel.R` depth limits:

- **S/M/L vessels** (`depth.lim = 5`): avg P(surface) at the 5 m depth threshold
- **XL vessels** (`depth.lim = 15`): avg P(surface) at the 15 m depth threshold

```{r}
#| label: diel-scenario-risk

# extract average P(surface) by depth from digitized table
# depth limits per SEFSC/VesselStrikeRiskModel Step1_RiskModel.R lines 86-91:
#   M/L: depth.lim = 5;  XL: depth.lim = 15
p_5m  <- p_surface |> filter(depth_m ==  5) |> select(avg_day, avg_night)
p_15m <- p_surface |> filter(depth_m == 15) |> select(avg_day, avg_night)

# assign P(surface) by vessel size and diel period
p_surface_lookup <- tibble::tibble(
  vessel_size = c("S", "M", "L", "XL", "S", "M", "L", "XL"),
  period      = rep(c("day", "night"), each = 4),
  p_surface   = c(
    # daytime: S/M/L use 5m threshold, XL uses 15m threshold
    rep(p_5m$avg_day, 3),  p_15m$avg_day,
    # nighttime: S/M/L use 5m threshold, XL uses 15m threshold
    rep(p_5m$avg_night, 3), p_15m$avg_night
  )
)

# compute diel-adjusted Garrison risk per area
# (multiply existing Garrison risk by P(surface) for each scenario)
garrison_diel <- garrison_risk |>
  inner_join(area_long, by = "cell_id", relationship = "many-to-many")

# day scenario
garrison_day <- garrison_diel |>
  left_join(
    p_surface_lookup |> filter(period == "day") |> select(-period),
    by = "vessel_size") |>
  mutate(
    risk_day          = garrison_risk * p_surface,
    risk_day_slowdown = garrison_slowdown_risk * p_surface)

# night scenario
garrison_night <- garrison_diel |>
  left_join(
    p_surface_lookup |> filter(period == "night") |> select(-period),
    by = "vessel_size") |>
  mutate(
    risk_night          = garrison_risk * p_surface,
    risk_night_slowdown = garrison_slowdown_risk * p_surface)

# summarize by area
diel_by_area <- garrison_day |>
  group_by(area_label) |>
  summarize(
    garrison_risk = sum(garrison_risk, na.rm = TRUE),
    risk_day      = sum(risk_day, na.rm = TRUE),
    .groups       = "drop") |>
  inner_join(
    garrison_night |>
      group_by(area_label) |>
      summarize(
        risk_night = sum(risk_night, na.rm = TRUE),
        .groups    = "drop"),
    by = "area_label") |>
  mutate(
    night_day_ratio = risk_night / risk_day,
    area_label      = factor(area_label, levels = area_order)) |>
  arrange(area_label)

diel_by_area |>
  kable(
    digits    = c(0, 4, 4, 4, 1),
    col.names = c(
      "Area", "Garrison Risk (P=1)",
      "Day Scenario", "Night Scenario", "Night/Day Ratio"),
    format.args = list(big.mark = ","),
    caption = "Diel risk scenarios: Garrison risk scaled by P(surface) from Kok et al. (2023).")
```

```{r}
#| label: diel-slowdown-comparison

# show day vs night with slowdown
diel_slowdown <- garrison_day |>
  group_by(area_label) |>
  summarize(
    risk_day          = sum(risk_day, na.rm = TRUE),
    risk_day_slow     = sum(risk_day_slowdown, na.rm = TRUE),
    .groups           = "drop") |>
  inner_join(
    garrison_night |>
      group_by(area_label) |>
      summarize(
        risk_night      = sum(risk_night, na.rm = TRUE),
        risk_night_slow = sum(risk_night_slowdown, na.rm = TRUE),
        .groups         = "drop"),
    by = "area_label") |>
  mutate(
    day_reduction_pct   = (risk_day - risk_day_slow) / risk_day * 100,
    night_reduction_pct = (risk_night - risk_night_slow) / risk_night * 100,
    area_label          = factor(area_label, levels = area_order)) |>
  arrange(area_label)

diel_slowdown |>
  select(area_label, risk_day, risk_day_slow, day_reduction_pct,
         risk_night, risk_night_slow, night_reduction_pct) |>
  kable(
    digits    = c(0, 4, 4, 1, 4, 4, 1),
    col.names = c(
      "Area",
      "Day Risk", "Day Slowdown", "Day Reduct. %",
      "Night Risk", "Night Slowdown", "Night Reduct. %"),
    format.args = list(big.mark = ","),
    caption = "Slowdown risk reduction under day vs. night scenarios.")
```

::: {.callout-warning}
## Caveats for Diel Scenario Analysis

- **n = 2 whales:** P(surface) values are digitized from @kokDiveBehaviorRicesWhale2023 Fig. 5 for Milky Way (2015) and Edna (2018), then averaged. Population-level dive behavior may differ from these two individuals.
- **Digitization method:** Cumulative % time curves were extracted using [WebPlotDigitizer](https://automeris.io/WebPlotDigitizer). The 30 m depth row is extrapolated from the 15-to-20 m per-meter rate (Fig. 5 only extends to 20 m).
- **Day/night traffic split unknown:** We apply P(surface) uniformly to all traffic. In reality, the proportion of vessel transits occurring during day vs. night varies by area and vessel type. General maritime patterns suggest most commercial traffic operates around the clock, but recreational traffic concentrates during daylight hours.
- **Assumed depth thresholds:** The strike zone depths (5 m for S/M/L, 15 m for XL) follow the SEFSC `Step1_RiskModel.R` depth limits and may not exactly represent the hazardous zone for all vessel configurations.
- **Conservative framing:** The night/day ratio quantifies how much *more* risk nighttime transits pose due to whale surfacing behavior. This has direct management relevance for potential nighttime speed restrictions or routing measures, even without precise vessel-level time-of-day data.
- **Additional tags pending:** NMFS tagged 7 Rice's whales in 2024; those dive profiles, when available, will substantially improve P(surface) estimates and reduce the uncertainty inherent in n = 2.
:::

## Deliverable Maps

### D1a: Rice's Whale Densities (Annual Average)

Map of annual average whale density (n = number of individuals per cell) from @litzCetaceanSeaTurtle2022.

```{r}
#| label: map-whale-density-annual
#| fig-cap: "Annual average Rice's whale density from @litzCetaceanSeaTurtle2022 model. Values represent number of individuals per 0.1° cell."

# get annual average whale density raster
r_whale_annual <- get_whale_raster(
  con        = con,
  month_int  = 0,
  metric     = "n",
  r_template = r_template
)

# create map with critical habitat overlay
map_rast_simple(
  r             = r_whale_annual,
  legend_title  = "Whale Density<br><small>(n per cell)</small>",
  colors        = "Spectral",
  group         = "whale_density",
  overlay_sf    = ply_crithab,
  overlay_color = "blue",
  overlay_label = "Critical Habitat"
)
```

### D1a-monthly: Monthly Whale Densities

```{r}
#| label: map-whale-density-monthly
#| results: asis
#| fig-cap: "Monthly Rice's whale density maps showing seasonal variation."

# create monthly maps as tabs
cat("::: {.panel-tabset}\n\n")

for (m in 1:12) { # m = 1
  cat(glue("#### {month.abb[m]}\n\n"))

  r_month <- get_whale_raster(
    con        = con,
    month_int  = m,
    metric     = "n",
    r_template = r_template
  )

  map_out <- map_rast_simple(
    r            = r_month,
    legend_title = glue("Whale Density<br><small>({month.abb[m]})</small>"),
    colors       = "Spectral",
    group        = glue("whale_{month.abb[m]}")
  )

  # print map
  print(htmltools::tagList(map_out))
  cat("\n\n")
}

cat(":::\n")
```

### D1a-cv: Whale Density Uncertainty (CV)

Map of coefficient of variation (CV) for annual average whale density, showing spatial uncertainty in the model predictions.

```{r}
#| label: map-whale-cv
#| fig-cap: "Coefficient of variation (CV) for annual average whale density estimates."

# get annual average CV raster
r_whale_cv <- get_whale_raster(
  con        = con,
  month_int  = 0,
  metric     = "cv",
  r_template = r_template
)

map_rast_simple(
  r            = r_whale_cv,
  legend_title = "CV<br><small>(uncertainty)</small>",
  colors       = "YlOrRd",
  group        = "whale_cv"
)
```

### D1a-boundaries: Study Area Boundaries

```{r}
#| label: map-boundaries

# basic leaflet map of study area boundaries
leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addPolygons(
    data        = ply_goa,
    fillColor   = "transparent",
    color       = "gray",
    weight      = 1,
    label       = "Study Area"
  ) |>
  addPolygons(
    data        = ply_crithab,
    fillColor   = "blue",
    fillOpacity = 0.2,
    color       = "blue",
    weight      = 2,
    label       = "Critical Habitat"
  ) |>
  addPolygons(
    data        = ply_subareas,
    fillColor   = "orange",
    fillOpacity = 0.3,
    color       = "orange",
    weight      = 2,
    label       = ~as.character(1:nrow(ply_subareas))
  )
```

### D1b: Vessel Traffic

Map of total vessel traffic (distance traveled in km) per 0.1° cell from 2023 AIS data.

```{r}
#| label: map-traffic
#| fig-cap: "Vessel traffic density (total km traveled) from 2023 AIS data."

r_traffic <- get_traffic_raster(
  con        = con,
  metric     = "km",
  r_template = r_template)

map_rast_simple(
  r             = r_traffic,
  legend_title  = "Traffic<br><small>(km)</small>",
  colors        = "YlOrRd",
  group         = "traffic",
  overlay_sf    = ply_crithab,
  overlay_color = "blue",
  overlay_label = "Critical Habitat")
```
### D1c: Strike Risk

Map of speed-weighted ship-strike risk per cell (whale density × traffic × Conn/Silber risk factor).

```{r}
#| label: map-risk
#| fig-cap: "Speed-weighted ship-strike risk for Rice's whale. Risk = whale density × vessel km × @connVesselSpeedRestrictions2013 speed factor."

# aggregate risk to per-cell totals (speed-weighted)
risk_per_cell <- risk_data |>
  mutate(
    risk_factor = get_risk_factor(speed_bin),
    cell_risk   = whale_density * traffic * risk_factor) |>
  group_by(cell_id) |>
  summarize(risk = sum(cell_risk, na.rm = TRUE), .groups = "drop")

r_risk <- terra::init(r_template, NA)
r_risk[risk_per_cell$cell_id] <- risk_per_cell$risk
names(r_risk) <- "strike_risk"

map_rast_simple(
  r             = r_risk,
  legend_title  = "Strike Risk<br><small>(weighted)</small>",
  colors        = "YlOrRd",
  group         = "strike_risk",
  overlay_sf    = ply_crithab,
  overlay_color = "blue",
  overlay_label = "Critical Habitat")
```

## Deliverable Tables

### D2: Whale Densities by Area

```{r}
#| label: tbl-whale-density

# summarize whale densities by area
whale_by_area <- whale_annual |>
  inner_join(area_long, by = "cell_id", relationship = "many-to-many") |>
  group_by(area_label) |>
  summarize(
    n_cells       = n(),
    mean_density  = mean(whale_density, na.rm = TRUE),
    total_density = sum(whale_density, na.rm = TRUE),
    .groups       = "drop") |>
  mutate(area_label = factor(area_label,
    levels = c("Study Area", "Critical Habitat",
               paste("Subarea", 1:7)))) |>
  arrange(area_label)

whale_by_area |>
  kable(
    digits    = 4,
    col.names = c("Area", "Cells", "Mean Density", "Total Density"))
```

### D3: Risk by Area (Unweighted)

Base risk = whale density × traffic distance (no speed weighting).

```{r}
#| label: tbl-risk-unweighted

risk_by_area |>
  calc_area_contribution(risk_col = "total_base_risk") |>
  select(area_label, total_base_risk, contribution_pct) |>
  kable(
    digits    = c(0, 2, 1),
    col.names = c("Area", "Base Risk", "% of Total"),
    format.args = list(big.mark = ","))
```

### D4: Risk by Area (Speed-Weighted)

Speed-weighted risk using @connVesselSpeedRestrictions2013 factors.

```{r}
#| label: tbl-risk-weighted

risk_by_area |>
  calc_area_contribution(risk_col = "total_speed_weighted_risk") |>
  select(area_label, total_speed_weighted_risk, contribution_pct) |>
  kable(
    digits    = c(0, 2, 1),
    col.names = c("Area", "Speed-Weighted Risk", "% of Total"),
    format.args = list(big.mark = ","))
```

### D5: Risk Reduction Scenarios

Risk reduction if all vessels slowed to ≤10 kt (baseline Conn/Silber factor = 1.0).

```{r}
#| label: tbl-risk-reduction

risk_by_area |>
  select(area_label, total_speed_weighted_risk, total_slowdown_risk,
         risk_reduction_abs, risk_reduction_pct) |>
  kable(
    digits    = c(0, 2, 2, 2, 1),
    col.names = c("Area", "Current Risk", "Slowdown Risk",
                  "Reduction", "Reduction %"),
    format.args = list(big.mark = ","))
```

### D7: Company-Specific Results

Per-company risk and potential reduction from fleet-wide slowdown to ≤10 kt.

```{r}
#| label: tbl-company-risk

# query traffic by cell, speed bin, and company
company_traffic_sql <- glue("
  SELECT
    sc.cell_id,
    co.company,
    {get_speed_bin_sql()} AS speed_bin,
    SUM(sc.dist_travelled_km) AS total_km
  FROM ship_cell sc
  JOIN ship s ON sc.mmsi = s.mmsi
  JOIN company co ON s.operator ILIKE '%' || co.operator_lookup || '%'
  WHERE sc.avg_speed_knots IS NOT NULL
  GROUP BY sc.cell_id, co.company, speed_bin
")

company_traffic <- DBI::dbGetQuery(con, company_traffic_sql)

# join with whale density and compute risk
company_risk_data <- company_traffic |>
  inner_join(whale_annual, by = "cell_id") |>
  mutate(
    traffic   = total_km,
    speed_bin = factor(speed_bin, levels = RISK_FACTORS$labels))

# summarize by company
company_risk <- summarize_risk_by_area(
  company_risk_data,
  area_col = "company",
  verbose  = FALSE)

company_risk |>
  arrange(desc(total_speed_weighted_risk)) |>
  kable(
    digits    = c(0, 2, 2, 2, 2, 1),
    col.names = c("Company", "Base Risk", "Speed-Weighted Risk",
                  "Slowdown Risk", "Reduction", "Reduction %"),
    format.args = list(big.mark = ","))
```

## Cleanup

```{r}
#| label: cleanup

DBI::dbDisconnect(con, shutdown = TRUE)
```

## References

::: {#refs}
:::
